The antigenic landscape of human influenza N2 neuraminidases from 2009 until 2017

Version 1

Abstract

Abstract

Human H3N2 influenza viruses are subject to rapid antigenic evolution which translates into frequent updates of the composition of seasonal influenza vaccines. Despite these updates, the effectiveness of influenza vaccines against H3N2-associated disease is suboptimal. Seasonal influenza vaccines primarily induce hemagglutinin-specific antibody responses. However, antibodies directed against influenza neuraminidase (NA) also contribute to protection. Here, we analyzed the antigenic diversity of a panel of N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. The antigenic breadth of these NAs was determined based on the NA inhibition (NAI) of a broad panel of ferret and mouse immune sera that were raised by infection and recombinant N2 NA immunization. This assessment allowed us to distinguish at least 4 antigenic groups in the N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. Computational analysis further revealed that the amino acid residues in N2 NA that have a major impact on susceptibility to NAI by immune sera are in proximity of the catalytic site. Finally, a machine learning method was developed that allowed to accurately predict the impact of mutations that are present in our N2 NA panel on NAI. These findings have important implications for the renewed interest to develop improved influenza vaccines based on the inclusion of a protective NA antigen formulation.

Introduction

Neuraminidase (NA) and hemagglutinin (HA) constitute the two major antigenic proteins at the surface of influenza A and B virions (Uyeki et al., 2022). While HA mediates binding to sialic acids and membrane fusion, NA catalyses the removal of sialic acid residues from sialylated glycoproteins and glycolipids. This complementary role of HA and NA enables influenza virions to penetrate the mucus barrier that lines the respiratory epithelium and newly formed virions to be released from the infected cell (de Vries et al., 2020; Matrosovich et al., 2004; Palese et al., 1974). Immunity against influenza is associated with the presence of anti-HA antibodies in serum, as determined by a hemagglutination inhibition assay. In line with this, the composition of licensed influenza vaccines is primarily focused on the antigenic characteristics of HA as the protective vaccine antigen (Cox, 2013). Multiple studies, however, have underscored the protective potential of anti-NA immune responses as determined by NA inhibition (NAI) assays (Couch et al., 2013; Maier et al., 2019; Memoli et al., 2016; Monto and Kendal, 1973; Monto et al., 2015; Ng et al., 2019; Weiss et al., 2019). There is also evidence that pre-existing anti-N2 immunity induced by prior exposure to H2N2 viruses that circulated in the population from 1957 to 1968, tempered the impact of the H3N2 pandemic virus that emerged in 1968 (Eickhoff and Meiklejohn, 1969; Jester et al., 2020; Monto and Kendal, 1973; Viboud et al., 2005). Further, the pharmacological inhibition of NA activity is a well-established therapeutic strategy to try to reduce influenza severity and duration (Leneva et al., 2000; Yen et al., 2005). In addition, antibodies with NAI activity can protect against influenza virus infection in pre-clinical models (Chen et al., 2018; Deroo et al., 1996; Rosu et al., 2022; Wan et al., 2018; Wohlbold et al., 2017).

Since its appearance in 1968, human H3N2 viruses have been evolving faster than the co-circulating H1N1 and influenza B viruses (Bedford et al., 2014; Potter et al., 2019). The overall Influenza virus vaccine effectiveness varies between seasons and, despite frequent updates of the H3N2 HA component, the effectiveness of influenza vaccines against H3N2-associated disease ranges between 5 and 39%, which is lower than the effectiveness against H1N1 and influenza B (Belongia et al., 2016). In this context the use of NA as a complementary vaccine antigen might improve influenza vaccine effectiveness, for example by compensating for HA-based antigenic mismatches in circulating influenza virus strains (Krammer et al., 2018).

Phylogenetic analyses of HA and NA genes of human H3N2 viruses revealed the asynchronous evolution of the 2 proteins. In addition, the evolutionary rate as determined by the accumulation of mutations over time, is estimated to be 1.5 times lower for NA than for HA (Westgeest et al., 2012). Despite the strong evidence that immune responses directed against NA can contribute to protection against influenza, a comprehensive insight in the antigenic breadth and evolution of the H3N2 NA antigen is lacking (Nachbagauer et al., 2017; Rajendran et al., 2017; Smith et al., 2004; Westgeest et al., 2012).

Here, we determined the antigenic landscape of N2 NAs derived from human H3N2 strains that were isolated between 2009 and 2017. This analysis allowed the identification of at least 4 N2 NA antigenic groups and to estimate the weight of amino acid changes on their contribution to NA antigenic drift.

Results

Generation of H3N2-derived NA immune sera in ferrets and mice

Our aim was to determine the antigenic relationship between NAs derived from human H3N2 viruses that circulated during nearly one decade. We first selected 44 H3N2 neuraminidase protein sequences obtained via GISAID, derived from viruses that were isolated between 2009 and 2017 (Elbe and Buckland-Merrett, 2017). Hierarchical clustering with minimax linkage was employed to ensure the set was representative of overall protein sequence diversity (Bien and Tibshirani, 2011). Next, 44 recombinant H1N2 viruses were generated that carried the internal and HA genes from PR8 virus and the NA genes of the selected H3N2 viruses. In addition, the corresponding tetrabrachion-stabilized recombinant N2 NA proteins were produced in Chinese hamster ovary (CHO) cells and purified, and are referred to as NACHOs (Catani et al., 2022). All purified NACHOs were tetrameric based on SEC-MALS analysis. The enzymatic activity of the NACHOs was determined based on the hydrolysis of the fluorogenic small molecule substrate 4-methylumbelliferyl-N-acetylneuraminic acid (MUNANA). The NACHOs were active with a specific activity ranging from 0.27 to 27.48 nmole/min/µg, indicating correct folding and assembly of the NA head domains (Table S1).

Immune sera against all 44 H3N2-derived NAs were generated in pairs of ferrets by a primary infection with the respective H1N2 reassortant viruses. This prime by infection was used to avoid the use of an adjuvant upon subsequent boost with the corresponding NACHOs. On day 28 after primary infection, all animals were boosted by intramuscular injection of homologous NACHO, except for the ferrets that had been primed by infection with A/Stockholm/15/2014 and A/Estonia/91621/2015, which were re-inoculated with the same respective H1N2 reassortant viruses. These two pairs of ferrets were not boosted with homologous NACHO because the corresponding NACHOs were not available at the time of the experiment. Blood samples were taken from the ferrets on days 14 and 42 after (primary) infection (Figure 1a). Priming by H1N2 infection resulted in seroconversion against the respective immobilized homologous NACHOs (Figure 1b). Subsequent immunization with recombinant NACHOs boosted the NACHO-specific IgG titers in day 42 serum samples. In contrast, serum samples from the ferrets that had been re-inoculated with H1N2 reassortant carrying NA derived from A/Stockholm/15/2014 or A/Estonia/91621/2015, showed no further increase in titers, presumably because prior exposure with the same H1N2 virus had induced a sterilizing immune response (Figure 1b).

Generation of anti-N2 sera panel from ferrets and mice.

(a) Ferret immunization and blood sampling scheme. The second immunization on day 28 was performed with homologous NACHO except for the ferrets that had been primed with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014, which were infected again on day 28 post prime with the same H1N2 viruses. (b) Homologous NACHO ELISA titer of ferret sera sampled on day 14 days after inoculation with H1N2 virus and 2 weeks after the boost with recombinant NA or by re-inoculation with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014 (indicated by arrows on the X-axis). (c) BALB/c immunization and blood sampling scheme. (d) Homologous NACHO serum IgG titers in pooled sera from mice that were immunized twice with recombinant NA. Mouse and ferrets representation were created with BioRender.com (2023).

We also generated NA immune sera in BALB/c mice. Groups of 5 mice were immunized two times, with a two-week interval, by intramuscular injection with 1 µg of the respective 43 NACHOs adjuvanted with AF03 (the NACHO A/West-Virginia/17/2012 was not available for mice immunization and ELISAs). Blood was collected from the immunized mice and serum prepared on day 28 after the prime immunization (Figure 1c). NACHO-specific IgG titers in the mouse immune sera, pooled from each group, ranged from 1:8100 to 1:72900 (Figure 1d).

The antigenic profile of N2 neuraminidase

Next, NAI titers were determined in the ferret sera using a panel of H6N2 reassortant viruses in an enzyme-linked lectin assay (ELLA), avoiding interference by anti-H1 HA antibodies in this assay. We generated 27 H6N2 influenza viruses with H6 HA derived from A/mallard/Sweden/2002 and NA derived from a subset of the 44 H3N2 viruses, thereby retaining NA sequence diversity and year of H3N2 virus isolation (Table S2). The 27 H6N2 viruses could be classified into 4 groups when ordered according to the phylogenetic relatedness of their NA sequence (Figure 2). The conversion rate of MUNANA into 4-methylumbelliferyl and acetylneuraminic acid by these H6N2 viruses revealed a KM that ranged from 20.7 to 144.8 µM as determined by the Michaelis-Menten equation (Figure 2 – figure supplement 1a and table S2). Interestingly, the KM differs between the distinct phylogenetic groups (Figure 2 – figure supplement 1a). There was a negative correlation between the KM values and the dilution of the H6N2 viruses required to set the 70% maximal NA activity in ELLA (Figure 2 – figure supplement 1b and table S2).

Breadth of N2 NA inhibition.

Phylogenetic tree based on the N2 NA head domain amino acid sequences and heat-map representing the average of normalized neuraminidase inhibition per H6N2 [log2 (max NAI/NAI)] determined in ferret sera after the boost (listed vertically). The red-to-blue scale indicates high-to-low NAI observed in ELLA against the H6N2 reassortants (listed at the bottom). UPGMA clustering of H6N2s inhibition profiles are shown on top of the heat map and coloured according to the phylogenetic groups.

Using this panel of 27 H6N2s, we determined the NAI titers in the sera obtained from the 44 pairs of ferrets 2 weeks after the boost. ELLAs were performed in duplicate for all serum samples against one H6N2 target at the same time. The reciprocal NAI titers of the ferret sera paired with H6N2 viruses are shown in table S3. In general, the NAI titers of the ferret sera were the highest against homologous and phylogenetically closely related H6N2 viruses.

In Figure 2, all the tested ferret sera are represented in the rows and the 27 H6N2 viruses against which the sera were evaluated in ELLA are presented in the columns. The serum samples were ordered according to the phylogenetic relatedness of the N2 NA head domain amino acid sequences, showing that these cluster in 4 groups (G1-G4). The ELLA-based IC50 NAI titers were plotted relative to the maximum NAI titer obtained against the respective H6N2 target viruses. The heatmap thus represents, for each H6N2 virus, the normalized ELLA IC50 titer expressed as log2 of the fold-difference relative to the serum with the highest IC50 value against that H6N2 virus (dark red squares indicate high and blue squares low IC50 values) (Figure 2).

Next, these normalized NAI values were clustered against each of the H6N2s using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA), which is a distance-based method. This clustering allowed to discern at least 4 major antigenic clusters (AC1 to -4) that are depicted at the top of the heat map (Figure 2). It is interesting to note that the H6N2 viruses with NA antigens derived from A/Wisconsin/16/15, A/Alaska/251/15, and A/Kansas/14/2017 do not segregate into AC2, as would be predicted based on the phylogenetic relatedness, but cluster in AC1 based on the UPGMA. However, the overall concordance of the antigenicity with the phylogenetic relatedness, based on the amino acid sequences of the NA head domains, was confirmed by hierarchical clustering with bootstrapping and k-means (Figure 2 – figure supplements 2a and 2b). In line with this, a positive correlation was observed between the pairwise phylogenetic distances and the normalized NAI IC50 values for each of the tested H6N2s (Figure 2 – figure supplement 3).

We further tested the NAI activity of ferrets’ sera obtained 2 weeks after the prime infection against at least one H6N2 virus from each antigenic cluster. These H6N2 viruses were selected based on their antigenic relatedness as deduced from the ferret immune sera analysis (Figure 2) with 2 representatives from group 1 (Tex12, and Per09) and 1 each from group 2 (Kan17), group 3 (Hel823), and group 4 (Ind11). The normalized data confirmed the antigenic clustering observed in sera after boost (Figure 2 – figure supplement 4).

The NAI titers were also determined in the sera obtained from the 43 groups of NACHO-immunized mice against 7 of the H6N2 viruses. These H6N2 viruses were selected as described above with the addition of HK14 and Swi13 H6N2s, both belonging to antigenic cluster 1. Serum NAI IC50 values against a given H6N2 virus were normalized against the IC50 of the mouse serum that had the highest NAI titers against the tested H6N2 virus. UPGMA-based clustering showed that the mouse NACHO immune sera displayed a similar recognition pattern as the ferret immune sera that had been generated by H1N2 priming followed by a NACHO boost (Figure 2). For example, HK14 H6N2 NA activity is strongly inhibited by sera from group 1, is poorly inhibited by most of the sera from group 2 and is hardly inhibited by group 3 and 4 sera from both ferrets and mice. Ind11 H6N2 NA activity, on the other hand, was inhibited by group 1 and 4, and hardly by group 2 and 3 immune sera. Overall, the normalized NAI IC50 values of the ferret and mouse immune sera correlated very well with coefficients values ranging from 0.33 to 0.78 (Figure 2 – figure supplement 5b).

Association analysis reveals amino acid residue substitutions that strongly impact NAI

NA of human influenza viruses is subject to antigenic drift, predominantly driven by the NAI activity that is detectable in the host serum (Air et al., 1985; Yasuhara et al., 2019). There are a total of 102 amino acid differences between the head domains of the NAs that were used to generate the 44 ferret immune sera (Figure 3 - Figure supplement 1). These 102 variable amino acids in the N2 head domain are predominantly exposed on the NA surface (Figure 3 – figure supplement 2). To try to identify which of these amino acid substitutions have a significant impact on NAI, we performed an association analysis for each of the tested H6N2 viruses. For this analysis, the NAI IC50 value was compared between groups of N2 sera according to the presence or absence of a substitution of any of the 102 amino acid position that are different between the panel of 44 tested N2s against which sera had been raised in ferrets. As a result, a p-value is obtained for each position that reflects its possible impact on NAI (Figure 3a). The significance values higher than 1.3 (p<0.05) of representative strains of the 4 antigenic groups were represented on the surface of N2 structure (Figure 3b). No obvious antigenic patches, however, could be distinguished on the N2 NA surface.

Association analysis reveals impact of amino acid mutations on NAI.

(a) Association study performed for each H6N2 virus (listed at the bottom) to determine the probability of each variable amino acid (listed on the right) present in the panel of 44 NAs to impact the NAI. The red-to-blue scale represents significance values (S=-log10P) associated with each amino acid substitution. The dark red in the color scale represent all values that are equal or higher than 3 (corresponding to p-values ≤ 0.001). The colour shades on the bottom side indicate the distinct phylogenetic groups. (b) The significant values (p<0.05) were represented in the surface of N2 tetramer (PDB accession number 4H53) for one representative strain of each phylogenetic group with a yellow to red scale of significance (-log10P).

A similar association analysis was performed using the NAI values obtained with the mouse NACHO immune sera. The calculated p-values obtained from the mouse and ferret NAI data were in line with each other (Figure 3 – figure supplement 3).

Residues close to the catalytic pocket are more likely to impact NAI

We next calculated the distance of every Cα atom of each amino acid present in the N2 structure (PDB: 6BR6, from A/Perth/16/2009 NA) to each Cα of the variable amino acids. We then determined the correlation of those distances with the p-values that were previously determined in the association analysis. The resulting correlation coefficients (Tau values), indicate a similar pattern among all 27 tested H6N2s (Figure 4a). When visualized on the N2 surface structure, the Tau values revealed a geometrical distribution with the negative correlations being in closer proximity to the catalytic pocket (Figure 4b). Similar data was obtained when mice sera were analyzed this way (Figure 4 – figure supplement 1).

Residues close to the catalytic pocket are more likely to impact NAI.

(a) Significance values (102 variable amino acids) obtained from the association study were correlated to the distance between every amino acid and each of the variable amino acids in the NA structure. The red-to-blue scale represents the Tau values obtained in the Kendall correlation. Amino acids 83 to 469 are shown from top to bottom in the heatmap. The H6N2s used in the NAI panel are listed at the bottom. The grey column (catalytic pocket (CP)) indicates the residues that are part of the catalytic pocket that directly interact with substrate (R118, D151, R152, R224, E276, R292, R371, and Y406) and the residues that play a key role in stabilizing the catalytic site (E119, R156, W178, S179, D198, I222, E227, H274, E277, N294, and E425). (b) Tau values obtained from Tex12 H6N2 correlation are represented on the surface of the N2 structure of A/RI/5+/1957 (PBD 4H53).

Prediction of the N2 antigenicity based on a machine learning method

The antigenic drift of human H3 HA has been extensively studied and modelled. (Meyer and Wilke, 2015; Smith et al., 2004; Westgeest et al., 2012; Wu et al., 2020). To model the NA drift observed in our panel of N2 NAs, we first calculated the antigenic distances between the distinct NA sera, using a similar approach as has been used to visualize the HA antigenic map (Cai et al., 2012). These antigenic distances can be visualized using classical multidimensional scaling, and the resulting antigenic map (Figure 5 – figure supplement 1a) revealed a distribution that is compatible with the representation of phylogenetic distances (Figure 5 – figure supplement 1b).

Then, using the antigenic distances and a random forest method (RF), a model was built based on amino acid sequence information. A similar model has been used by Yao et al. to predict the antigenicity of H3 HA (Yao et al., 2017). Here, the optimized model was built using a combination of 25 randomly sampled data in 1500 decision trees. The resulting model generates predictive antigenic distances that strongly correlate with the observed values (Figure 5a). The RF also enables scoring of amino acid residues according to their importance in the decision trees. One measure of importance can be defined as how early a change in a position is placed in the generated decision trees (mean minimal depth). Another measure of importance is how much the standard error mean increases when a variable is permuted. Finally, the prevalence of a variable amino acid in nodes of the decision tree can also indicate its importance. Those importance scores were represented in a multi-way plot (Figure 5b). The importance scores highlighted similar positions as previously determined in the association study. Finally, we used NA sequences of recent seasonal influenza vaccine recommended H3N2 strains to predict their antigenic distance relative to our panel of N2s, including 3 recent N2 NA components (from A/HongKong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021) that were not present in our data set. The predicted antigenic distances were plotted in the heatmap in a red to blue scale, indicating high to low antigenic distance, respectively (Figure 5c). The RF clearly reproduced the drift observed in AC2 versus AC1. The modelling also indicates that the antigenic distance of the 3 recent N2s to their corresponding group 2 NAs is increasing (represented by the dark blue turning light blue, from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G2) whereas their distance to group 4 NAs is decreasing (dark red becoming yellow from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G4).

Prediction of N2 antigenicity based on machine learning.

(a) Predicted vs observed normalized NAIs are shown. (b) Multi-way importance plot indicating the increase in mean standard error, the mean depth, and number of nodes. The top 10 most important amino acid changes are highlighted in blue and annotated with their position in the NA sequence. (c) The heatmap indicates the predicted antigenic distances of N2s from recent H3N2 isolates (at the bottom of the heatmap) relative to the N2s that were used in the NAI breadth panel (listed vertically). The antigenic distances were calculated based on NA protein sequence.

Discussion

The sequence diversity of N2 protein derived from the 44 selected human H3N2 viruses isolated between 2009 and 2017 allows to classify them in 4 major phylogenetic groups. Group 1 contains 22 N2 NAs, 4 of which were derived from vaccine recommended H3N2 strains: A/Perth/16/2009, A/Texas/50/2012, A/Switzerland/9715293/2013, and A/Hong Kong/4801/2014. Phylogenetic group 2 comprises 15 N2 NAs, one of which is derived from a vaccine H3N2 recommended strain: A/Singapore/INFIMH-16-0019/2016. The third phylogenetic group contains two NAs that are derived from H3N2 strains that had been isolated in Helsinki in 2013. Remarkably, the NA sequence of these two strains is closely related to that of H3N2s that circulated more than 15 years earlier (e.g. 99% identity with NA of A/Sidney/5/1997). The fourth phylogenetic group is composed of NAs derived from zoonotic cases with swine H3N2 strains isolated between 2012 and 2017. Swine-to-human transmission of H3N2 viruses have occurred multiple times in the USA in 2012 and continue to be detected at low frequency until to date (CDC, 2023; Elbe and Buckland-Merrett, 2017).

We assessed the antigenic diversity of the NA panel by generating 27 recombinant H6N2 viruses and 44 ferret NA immune sera produced by H1N2 priming followed by boosting with homologous recombinant NA (NACHO; n = 42) or by reinfection with H1N2 (A/Estonia/91621/2015 and A/Stockholm/15/2014). Most of the ferrets had detectable serum anti-N2 antibodies by day 14 after infection with H1N2. Anti-NA antibody titers increased after a boost with non-adjuvanted homologous recombinant CHO-produced soluble tetrameric NA suggesting that the use of recombinant NA may boost naturally acquired anti-NA antibody responses in the human population.

The NAI activity of the ferret sera against 27 tested H6N2s reassortants revealed a pattern of inhibition that largely coincides with the phylogenetic distances based on the NA head domain sequences. In the tested panel of H6N2s, at least 4 major antigenic groups were identified. We noted that H6N2 with Kan17 NA had a broader profile of inhibition than other H6N2 viruses from group 2, being also inhibited by serum raised against NA from phylogenetic groups 3 and, to a lesser extent, group 4. Additionally, the H6N2s Wis15 and Ala15 have an antigenic behaviour that is more closely related to antigenic cluster 1, despite belonging to group 2 according to their phylogenetic similarities. Based on the NAI panel, an association analysis was performed to try to identify amino acid changes that affect NAI. It is interesting to note that the amino acid substitutions S245N and S247T, combined with substitution P468H, were identified in this association analysis. Those substitutions were described as major contributors of the NA antigenic drift that occurred in H3N2 in 2014 (Wan et al., 2019). NAs from A/Wisconsin/16/2015 and A/Alaska/251/2015 also have the S245N and S247T substitutions and phylogenetically belong to group 2 NAs. However, those 2 NAs lack the P468H mutation and are antigenically closer to group 1 NAs, confirming the requirement of P468H combined with the S245N and S247T substitutions for this NA antigenic drift (Wan et al., 2019). This observation suggests that NA antigenic drift may sometimes require a combination of amino acid changes rather than single mutations. Wang and colleagues have very elegantly demonstrated that N2 antigenic evolution is constrained by charge balances in an antigenic region, which can partially explain the observation that NA amino acids can undergo co-evolution to result in antigenic drift (Wang et al., 2021). Interestingly, NAs originated from H3N2 variants classified in phylogenetic group 4 were antigenically closer to N2s belonging to phylogenetic group 1. This suggest that the antigenic drift of recent NAs from group 2 increases the antigenic distance from the N2s of swine origin.

Noteworthy, the analysis of the sera obtained from ferrets after infection shows that, despite the lower homologous anti-NA titers, the inhibition pattern evaluated against 5 distinct H6N2s (composed of at least one representative of each antigenic group) is in accordance with the response obtained after boosting with recombinant protein. This suggests that a boost with homologous recombinant soluble tetrameric NA does not alter the breadth of the antibody response induced by H1N2 infection only.

NAI was also determined against 7 distinct H6N2 reassortant viruses using a panel of 43 anti-N2 NA sera raised in mice by immunization with adjuvanted NACHOs. The observed NAI responses again allowed clustering of the N2 NAs into 4 main antigenic groups, which correlates with data obtained in ferrets. Studies in mice can thus offer a convenient alternative to the use of ferrets in future NA antigenicity studies. This data also supports the hypothesis that there may be universal rules of immunogen antigenicity and immunogenicity (Altman et al., 2015).

The determined KM values of the H6N2 viruses used here, indicates that substrate affinity tends to vary according to the phylogenetic relatedness of N2s. The N-glycosylation at position 245 was previously described to be responsible for reducing N2 affinity for its substrate (Powell and Pekosz, 2020). Unlike group 2, which is primarily characterized by the presence of a N-glycan at position 245, NAs from phylogenetic group 4 have a low KM, indicating an increased affinity for substrate. This result suggests that phylogenetic drift can be accompanied by changes in NA enzymatic properties. Curiously, the Ala15, Wis15, and Kan17, despite having antigenic properties closer to group 1, still have KM values that are comparable to those of group 2 N2 NAs.

After the determination of NA antigenic distances, we used random forest modelling to estimate the impact of amino acid changes on antigenic distances. We obtained a strong correlation of predicted versus observed antigenic distances. However, it is important to note that the training dataset was based on the 102 mutations present in our panel, limiting the predictions for other positions. Furthermore, random forest enabled the classification of the variable amino acids according to importance scores. It is important to highlight that co-occurring mutations in our panel (the ones present in main branches of the phylogenetic tree) cannot be individually assessed by association analysis or the random forest model. The individual weight of those mutations on NA drift remains to be experimentally demonstrated. Interestingly the modelling could put in evidence the gradual drift of recent N2s which were not present in the dataset (A/Hong Kong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021).

The protective potential of NA antigen against influenza has been underscored by multiple studies. It is likely that NA will be included as a standard component of next generation influenza vaccines. The difficulties of producing stable tetrameric NA may be overcome in the future thanks to protein design or with the advent of mRNA-based vaccines (Catani et al., 2022; Chivukula et al., 2021; Ellis et al., 2022). The antigenic modelling associated with sequence surveillance may increase our predictive ability, highlighting major potential drifts and support NA antigen designs that may elicit broader NAI responses compared to natural NA.

In summary, our study shows that immunization with recombinant N2 NA immunogen can boost anti-NA serum antibody responses in ferrets that were primed by infection with an influenza virus that carries the homologous NA. Prime by infection allowed to use non-adjuvanted NACHO in the subsequent boost. At least 4 major antigenic groups could be distinguished in N2 NA derived from human H3N2 viruses that were isolated between 2009 to 2017, and those antigenic groups predominantly correspond with the phylogenetic relatedness of the N2 head domains. In addition, the amino acid substitutions in close proximity of the NA catalytic pocket are more likely to affect NAI patterns.

Material and Methods

Rescue of H1N2 and H6N2 reassortants

All reassortant H1N2 and H6N2 viruses included in the study were generated by reverse genetics, with each reassortant expressing the targeted NA antigen, internal genes from A/Puerto Rico/8/1934 H1N1 (PR8), and the HA from PR8 H1N1 or A/mallard/Sweden/81/2002 H6N1, respectively. Briefly, HA and NA segments including non-coding regions were generated by custom gene synthesis (Geneart AG), while PR8 internal segments were derived from a viral isolate as previously published (Marsh and Tannock, 2005). All segments were cloned into a bi-directional transcription plasmid derived from pUC57 (Genscript) including polymerase (Pol) I and Pol II promoters, as previously described (Hoffmann et al., 2000), and the entire set of 8 plasmids was used to transfect 293FT cells (Thermo Fisher Scientific) using Lipofectamine 2000 CD (Thermo Fisher Scientific). Twenty-four hours after transfection, Madin-Darby Canine Kidney Cells (MDCK-) ATL cells (ATCC) were added to the transfected cells in the presence of TPCK-treated trypsin (Sigma) to allow influenza virus propagation. Cell culture supernatants containing influenza virus were harvested 7 days post-MDCK addition and blindly passaged in 8–10-day-old embryonated chicken eggs (Charles River Laboratories, Inc.). Inoculated eggs were incubated at 37°C for 48 h, then cooled to 4°C for 12 h, prior to allantoic fluid harvest and clarification by low-speed centrifugation (3,000 rpm, 20 min). For H6N2 viruses only, high-yield stocks were generated by an additional passage in eggs as described above. Virus titers were determined by plaque assay on MDCK cells.

Design and production of recombinant proteins

The coding information of the NACHO proteins was cloned under the transcriptional control of the CMV promoter in the pCDNA3.4 plasmid with a CD5 secretion signal, an amino-terminal His-tag and the thrombin cleavage signal followed by tetrabrachion-NA. NACHOs were expressed in a mammalian cell culture system as previously described with modifications (Catani et al., 2022).

Animal experiments

The ferret immunization experiment was conducted at Bioqual, Inc., MD (USA). The experiments were approved by the Institutional Animal Care and Use Committee of the Bioqual (reference number 17-021.6). Outbred 18-22 weeks old naïve male and female Fitch ferrets were obtained from Triple F, PA (USA) and housed under specified pathogen-free conditions with food and water ad libitum. Ferrets were primed on day 0 by intranasal challenge with influenza H1N2 reassortant virus (1,000µl/dose, split evenly between nostrils). Four weeks after initial infection (day 28), ferrets were immunized with 45 μg of NACHO (500µl/dose, intramuscularly, split evenly between hind thighs) corresponding to homologous NA used in day 0 prime. Alternatively, and due to low yield of selected recombinant NACHOs, ferrets were boosted on day 28 with 15 μg of homologous NACHO (A/Perth/16/2009nib-64, A/Tasmania/1018/2015, A/Utah/11/2011, and A/West Virginia/17/2012) or were inoculated with the same H1N2 reassortant virus used in the prime (A/Estonia/91621/2015 and A/Stockholm/15/2014). Sera from bleeds two weeks after the prime (day 14) and booster vaccination (day 42) were tested for immunogenicity and seroconversion in serological assays.

The mouse immunization experiments were conducted according to the Belgian legislation (Belgian Law 14/08/1986 and Belgium Royal Decree 06/04/2010) and European legislation on protection of animals used for scientific purposes (EU directives 2010/63/EU and 86/609/EEC). Experimental protocols were all approved by the Ethics Committee of the Vlaams Instituut voor Biotechnologie (VIB), Ghent University, Faculty of Science (EC2019-035). Female BALB/c mice, aged 6–7 weeks, were purchased from Charles River. The mice were housed in a specified pathogen-free animal house with food and water ad libitum. Animals were immunized intramuscularly in the right quadriceps. Immunization was performed with 50 µl containing 1 µg of recombinant protein (tetrabrachion stabilized NAs, referred to as NACHOs). All, except A/West Virginia/17/2012 NACHOs were used as immunogens in mice. All immunizations were adjuvanted with a 1:1 volume of AF03 (25 μl of antigen in PBS + 25 μl of AF03 per dose) (Caillet et al., 2010).

Two or three weeks after the boost immunization, mice were humanely sacrificed with an anaesthetic overdose of Nembutal (6 mg per mouse) and bled retro-orbitally. The obtained blood samples were incubated for 30 min at 37°C to allow clotting, which was followed by centrifugation at 10,500 g for 5 minutes. The supernatant (the serum) was recovered and submitted to a second centrifugation at 10,500 g for 5 minutes. Cleared sera were stored at -20°C before use in serological assays.

ELISA

Anti-tetNA IgG levels in mouse and ferret sera were determined by capture ELISA using tetNAs in PierceTM nickel-coated plates (cat. # 15442). Recombinant proteins were diluted in DPBS (Life technologies cat. # 14040-182) (tetNA was diluted to 0.5 µg/ml). Then, 50 µl of the coating antigen solutions was added to each well and the plate was incubated at room temperature for 1 h on a shaking platform. The wells of the plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and blocked for 1h with 1% BSA (for the mouse sera) or 10% milk (for the ferret sera) in DPBS. After blocking, the wells of the plates were washed once with PBS-T and incubated with a 3-fold serial dilution, starting at a 1/100 dilution, of serum in DPBS, 0.5% BSA, 0.05% tween-20 for 2h at room temperature on a shaking platform. The wells of the plates were then washed 5 times with PBS-T and incubated with a 1:5000 dilution of anti-mouse IgG-HRP (GE healthcare cat. # NA931-1ml), or with a 1:100000 dilution of anti-ferret IgG-HRP (Abcam cat. # 112770) in DPBS, 0.5% BSA, 0.05% Tween-20. The 3,3’,5,5’-tetramethylbenzidine (TMB) substrate (BD cat. # 555214) was added after 3 washes with PBS-T and the reaction was stopped after 5 min by addition of 50 µl of 1 M H2SO4. The optical density (O.D.) in each well was determined at 450 nm and, as a reference, 655 nm using an iMark™ Microplate Absorbance Reader (Bio-rad). The end point titer was determined for each serum sample by scoring the dilution that resulted in an O.D. that was equal to or 2 times higher than the background O.D. obtained from pre-immune control sera dilution series.

Enzyme-linked lectin assay to determine neuraminidase inhibition titers

Fetuin (Sigma cat. # F3385) was diluted into coating buffer (KPL cat. # 50-84-01) to a concentration of 25 µg/ml and 50 µl was added to the wells of Nunc MaxiSorp™ plates (Thermofisher cat. # 44-2404-21), which were incubated overnight at 4°C. The coated plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and incubated overnight with 60 μl of a dilution of H6N2 and 60 µl of 2-fold serial dilution of serum, starting at a 1/20 dilution, in sample buffer (1X MES VWR cat. # AAJ61979-AP: 20mM CaCl2, 1% BSA, 0.5% Tween-20). These dilutions of the H6Nx viruses correspond to the 70% maximum activity of NA from the respective viruses as determined in the ELLA assay. Fetuin-coated plates were then washed 3 times with PBS-T and incubated for 1 h with a solution of PNA-HRP (cat. # L6135-1MG, Sigma) at 5 μg/ml in conjugate diluent (MES pH 6.5, 20 mM CaCl2, 1% BSA). The plates were washed 3 times with PBS-T, TMB substrate was added and then the plates were incubated for 5 minutes before the reaction was stopped by the addition of 50 µl of 1 M H2SO4. The optical density was measured at 450 nm and, as a reference, 655 nm in an iMark™ Microplate Absorbance Reader (Bio-rad). Half maximum inhibitory concentrations (IC50) values were determined by non-linear regression analysis (GraphPad Prism software).

NA Enzyme Kinetics Assay

The enzymatic activity of NACHOs and H6N2s was determined with the fluorogenic small substrate 4-methylumbelliferyl)-α-d-N-acetylneuraminic acid (MUNANA) as previously described (Marathe et al., 2013). First H6N2 dilutions were tested to ensure that MUNANA conversion into 4-methylumbelliferone (4-MU) generated fluorescence over time at a linear ratio. Ten microliters of 1 mM of MUNANA (Sigma cat # M8639) in 200 mM sodium acetate buffer (pH 6.5) containing 2 mM CaCl2 and 1% butanol was incubated with 40 µl of PBS solution containing viral dilution. Conversion to 4-MU was monitored every 2 min for 1 h using a BMG Fluostar OPTIMA reader (excitation at 365 nm and emission determined at 450 nm). Second, using the selected dilution of virus, the assay was repeated with dilutions of MUNANA ranging from 2.5 to 100 µM. A standard curve with 800 to 25 pmoles/well of 4-MU (Sigma cat # M1508) was used to extrapolate the molar conversion of MUNANA. KM was calculated using GraphPad Prism version 8.4.3.

Association study

For the association analysis, under the assumption of homogeneous genetic relatedness among the selected H3N2 virus strains, a simple linear regression model for mutation-trait association analysis was fitted to the data for each H6N2 separately. The significance values were calculated by comparing IC50 values of sera that were classified into 2 groups according to the presence or absence of variation in the amino acid of reference Tex12 sequence.

Correlations with distances from the NA catalytic pocket

Tau values were determined by correlating distances with significance values obtained from the association analysis or the importance scores obtained from machine learning. A similar method was described by Meyer et al. to define geometrical constraints of HA evolution (Meyer and Wilke, 2015). Briefly, the relative distances of amino acids were calculated based on the structure of A/Perth/2009 (H3N2) NA (PDB: 6BR6) (Hadházi et al., 2018). The distances were calculated from each Cα to every other Cα. The distances from every residue to the 98 variable residues that are present in the panel of N2s and 6BR6 structure were then correlated with the significance values that were obtained in the association analysis, or importance scores obtained in the machine learning. Correlation coefficients were visualized on the structure of N2 NA derived from pandemic A/RI/5+/1957 H2N2 virus (PDB: 4H53)(Vavricka et al., 2013) using Chimera 1.14.

Random Forest algorithm

The Random Forest algorithm was adapted from (Yao et al., 2017). Briefly, the non-conserved positions in the multiple sequence alignment of N2 head domain amino acid sequences of all tested N2s were selected. The feature matrix (X) was created by assessing matches/mismatches on each non-conserved position for each pair of N2 head domain amino acid sequences as: where Xi represents the non-conserved positions in the multiple sequence alignment m is the number of non-conserved positions (102 in total)

Vi represents the N2 head domain sequence for each tested NA

n is the number of tested sera (44 in total)

0 represents a match, 1 represents a mismatch in the alignment

The response vector (Y) is composed of the antigenic distances for each pair of tested sera. These antigenic distances were determined based on NAI, similarly to the method used to determine HI based immunogenic distances that was described by Cai et al. In summary, the distances between sera was calculated as , which represent average distances between 2 serum samples among all tested H6N2s (Cai et al., 2012). We applied the random forest function from the randomForest package in R to construct the model. We set the bootstrapping (tree) number to 1500 and the number of features to 25 for each tree. Accuracy was determined as the root-mean-square error (RMSE), calculated using the RMSE function from the yardstick package in R.

Additionally, predicted antigenic distances of the out-of-bag samples were compared to the observed distances by fitting a linear model and calculating the R2value. To calculate the importance of each feature (Xi), the following metrics were calculated using the measure importance function of the randomForestExplainer package in R:

- mean minimal depth: mean of the depths of the node in each tree that splits on this feature and is the closest to the root of the tree (minimal depth). The lower, the more important the feature since a low depth means that many observations are divided into groups based on this feature.

- mse decrease: mean decrease of accuracy in predictions on out of bag samples when the feature is excluded from the model. The higher, the more important the feature.

- number of nodes: total number of nodes in the forest that use the feature for splitting. The higher, the more important the feature.

eLife Assessment

This study presents valuable data on the antigenic properties of neuraminidase proteins of human A/H3N2 influenza viruses sampled between 2009 and 2017. The antigenic properties are found to be generally concordant with genetic groups. However, the evidence for claims concerning the molecular basis of the antigenic differences remains incomplete, as the computational methods have been insufficiently validated.

Reviewer #1 (Public Review):

Summary

The authors investigated the antigenic diversity of recent (2009- 2017) A/H3N2 influenza neuraminidases (NAs), the second major antigenic protein after haemagglutinin. They used 27 viruses and 43 ferret sera and performed NA inhibition. This work was supported by a subset of mouse sera. Clustering analysis determined 4 antigenic clusters, mostly in concordance with the genetic groupings. Association analysis was used to estimate important amino acid positions, which were shown to be more likely close to the catalytic site. Antigenic distances were calculated and a random forest model was used to determine potential important sites.

This has the potential to be a very interesting piece of work. At present, there are inconsistencies in the methods, results and presentation that limit its impact. In particular, there are weaknesses in some of the computational work.

Strengths

1. The data cover recent NA evolution and a substantial number (43) of ferret (and mouse) sera were generated and titrated against 27 viruses. This is laborious experimental work and is the largest publicly available neuraminidase inhibition dataset that I am aware of. As such, it will prove a useful resource for the influenza community.

2. A variety of computational methods were used to analyse the data, which give a rounded picture of the antigenic and genetic relationships and link between sequence, structure and phenotype.

Weaknesses

1. Inconsistency in experimental methods

Two ferret sera were boosted with H1N2, while recombinant NA protein for the others. This, and the underlying reason, are clearly explained in the manuscript. The authors note that boosting with live virus did not increase titres. Nevertheless, these results are included in the analysis when it would be better to exclude them (Figure 2 shows much lower titres to their own group than other sera).

2. Inconsistency in experimental results

Clustering of the NA inhibition results identifies three viruses which do not cluster with their phylogenetic group. Again this is clearly pointed out in the paper. Further investigation of this inconsistency is required to determine whether this has a genetic basis or is an experimental issue. It is difficult to trust the remaining data while this issue is unresolved.

3. Inconsistency in group labelling

A/Hatay/4990/2016 & A/New Caledonia/23/2016 are in phylogenetic group 1 in Figure 2 and phylogenetic group 1 in Figure 5 - figure supplement 1 panel a.

A/Kansas/14/2017 is selected as a representative of antigenic group 2, when in Figure 2 it is labelled as AC1 (although Figure 2 - supplement 4 which the text is referring to shows data for A/Singapore/Infimh-16-0019/2016 as the representative of AC2). A/Kansas/14/2017 is coloured and labelled as AC2 in Figure 2 - supplement 5.

The colouring is changed for Figure 3a at the bottom. A/Heilongjiang-Xiangyang/1134/2011 is coloured the same as AC4 viruses when it is AC1 in Figure 2.

This lack of consistency makes the figures misleading.

4. Data not presented, without explanation

The paper states that 44 sera and 27 H6N2 viruses were used (line 158). However, the results for the Kansas/14/2017 sera do not appear to be presented in any of the figures (e.g. Figure 2 phylogenetic tree, Figure 5 - figure supplement 1). It is not obvious why these data were not presented. The exclusion of this serum could affect the results as often the homologous titre is the highest and several heatmaps show the fold down from the highest titre.

5. The cMDS plot does not have sufficient quality assurance

A cMDS plot is shown in Figure 5 - figure supplement 1, generated using classical MDS. The following support for the appropriateness of this visualisation is not given.

a. Goodness of fit of the cMDS projection, including per point and per titre.

b. Testing of the appropriate number of dimensions (the two sera from phylogenetic group 3 are clustered with phylogenetic group 2; additional dimensions might separate these groups).

c. A measure of uncertainty in positioning, e.g. bootstrapping.

d. A sensitivity analysis of the assumption about titres below the level of detection (i.e. that <20 = 10).

Without this information, it is difficult to judge if the projection is reliable.

6. Choice of antigenic distance measure

The measure of antigenic distance used here is the average difference between titres for two sera. This is dependent on which viruses have been included in the analysis and will be biased by the unbalanced number of viruses in the different clusters (12, 8, 2, 5).

7. Association analysis does not account for correlations

For each H6N2 virus and position, significance was calculated by comparing the titres between sera that did or did not have a change at that position. This does not take into account the correlations between positions. For haemagglutinin, it can be impossible to determine the true antigenic effects of such correlated substitutions with mutagenesis studies.

8. Random forest method

25 features are used to classify 43 sera, which seems high (p/3 is typical for classification). By only considering mismatches, rather than the specific amino acid changes, some signals may be lost (for example, at a given position, one amino acid change might be neutral while another has a large antigenic effect). Features may be highly, or perfectly correlated, which will give them a lower reported importance and skew the results.

Reviewer #2 (Public Review):

Summary:

The authors characterized the antigenicity of N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 using ferret and mice immune sera. Four antigenic groups were identified, which correlated with their respective phylogenic/ genetic groups. Among 102 amino acids differed by the 44 selected N2 proteins, the authors identified residues that differentiate the antigenicity of the four groups and constructed a machine-learning model that provides antigenic distance estimation. Three recent A(H3N2) vaccine strains were tested in the model but there was no experimental data to confirm the model prediction results.

Strengths:

This study used N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 and generated corresponding panels of ferret and mouse sera to react with the selected strains. The amount of experimental data for N2 antigenicity characterization is large enough for model building.

Weaknesses:

The main weakness is that the strategy of selecting 44 A(H3N2) viruses from 2009-2017 was not explained. It is not clear if they represent the overall genetic diversity of human A(H3N2) viruses circulating during this time. A comprehensive N2 phylogenetic tree of human A(H3N2) viruses from 2009-2017, with the selected 44 strains labeled in the tree, would be helpful to assess the representativeness of the strains included in the study. The second weakness is the use of double-immune ferret sera (post-infection plus immunization with recombinant NA protein) or mouse sera (immunized twice with recombinant NA protein) to characterize the antigenicity of the selected A(H3N2) viruses. Conventionally, NA antigenicity is characterized using ferret sera after a single infection. Repeated influenza exposure in ferrets has been shown to enhance antibody binding affinity and may affect the cross-reactivity to heterologous strains (PMID: 29672713). The increased cross-reactivity is supported by the NAI titers shown in Table S3, as many of the double immune ferret sera showed the highest reactivity not against its own homologous virus but to heterologous strains. Although the authors used the post-infection ferret sera to characterize 5 viruses (Figure 2, Figure Supplement 4), the patterns did not correlate well. If the authors repeat the NA antigenic analysis using the post-infection ferret sera with lower cross-reactivity, will the authors be able to identify more antigenic groups instead of 4 groups? Another weakness is that the authors used the newly constructed model to predict the antigenic distance of three recent A(H3N2) viruses but there is no experimental data to validate their prediction (eg. if these viruses are indeed antigenically deviating from group 2 strains as concluded by the authors).

Reviewer #3 (Public Review):

Summary:

This paper by Portela Catani et al examines the antigenic relationships (measured using monotypic ferret and mouse sera) across a panel of N2 genes from the past 14 years, along with the underlying sequence differences and phylogenetic relationships. This is a highly significant topic given the recent increased appreciation of the importance of NA as a vaccine target, and the relative lack of information about NA antigenic evolution compared with what is known about HA. Thus, these data will be of interest to those studying the antigenic evolution of influenza viruses. The methods used are generally quite sound, though there are a few addressable concerns that limit the confidence with which conclusions can be drawn from the data/analyses.

Strengths:

- The significance of the work, and the (general) soundness of the methods.

- Explicit comparison of results obtained with mouse and ferret sera.

Weaknesses:

- Approach for assessing the influence of individual polymorphisms on antigenicity does not account for the potential effects of epistasis.

- Machine learning analyses were neither experimentally validated nor shown to be better than simple, phylogenetic-based inference.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary

The authors investigated the antigenic diversity of recent (2009- 2017) A/H3N2 influenza neuraminidases (NAs), the second major antigenic protein after haemagglutinin. They used 27 viruses and 43 ferret sera and performed NA inhibition. This work was supported by a subset of mouse sera. Clustering analysis determined 4 antigenic clusters, mostly in concordance with the genetic groupings. Association analysis was used to estimate important amino acid positions, which were shown to be more likely close to the catalytic site. Antigenic distances were calculated and a random forest model was used to determine potential important sites.

This has the potential to be a very interesting piece of work. At present, there are inconsistencies in the methods, results and presentation that limit its impact. In particular, there are weaknesses in some of the computational work.

Strengths

(1) The data cover recent NA evolution and a substantial number (43) of ferret (and mouse) sera were generated and titrated against 27 viruses. This is laborious experimental work and is the largest publicly available neuraminidase inhibition dataset that I am aware of. As such, it will prove a useful resource for the influenza community.

(2) A variety of computational methods were used to analyse the data, which give a rounded picture of the antigenic and genetic relationships and link between sequence, structure and phenotype.

Weaknesses

(1) Inconsistency in experimental methods

Two ferret sera were boosted with H1N2, while recombinant NA protein for the others. This, and the underlying reason, are clearly explained in the manuscript. The authors note that boosting with live virus did not increase titres. Nevertheless, these results are included in the analysis when it would be better to exclude them (Figure 2 shows much lower titres to their own group than other sera).

As an exercise, we have excluded the H1N2 boosted ferrets sera and no major impact was observed in the antigenic grouping (see Author response image 1a). Another way to control for differences in immunogenicity is to normalize the NAI values with the homologous ELISA titers for each antigen. Clustering based on these ELISA normalized NAI titers reveals the same 4 distinct antigenic groups but with one change: Kan17 is shifted from group 1 to group 2 (Author response image 1b). Note that a homologous ELISA titer is not available for A/West-Virginia/17/2012 and thus this serum sample is not included in Author response image 1b.

Antigenic and phylogenetic relatedness of N2 NAs.

Phylogenetic tree based on the N2 NA head domain amino acid sequences and heat-map representing the average of normalized neuraminidase inhibition titer per H6N2 [log2 (max NAI/NAI)] determined in ferret sera after the boost (listed vertically). The red-to-blue scale indicates high-to-low NAI observed in ELLA against the H6N2 reassortants (listed at the bottom). UPGMA clustering of H6N2s inhibition profiles are shown on top of the heat map and colored according to the phylogenetic groups.(a) Based on the ferret sera with exclusion of the sera that were obtained following prime-boost by infection with H1N2 (A/Estonia/91625/2015 and A/Stockholm/15/2014). (b) Based on serum NAI titers that were normalized by the homologous ELISA titer.

(2) Inconsistency in experimental results

Clustering of the NA inhibition results identifies three viruses which do not cluster with their phylogenetic group. Again, this is clearly pointed out in the paper. Further investigation of this inconsistency is required to determine whether this has a genetic basis or is an experimental issue. It is difficult to trust the remaining data while this issue is unresolved.

We understand the concern of the reviewer. It is important to keep in mind that discrete grouping of antigens allows to visualize major antigenic drifts. However, within closely related groups the cross reactivity of antisera is more likely distributed in a spectrum. When we constructed an antigenic map based on the antigenic cartography algorithm (as described by Smith D. et al, 2004), Kansas17, Wis15, and Ala15 are positioned more closely to antigenic group 1 than the majority of other antigens that were classified as group 2 (Author response image 2a). Similar results were obtained when individual ferret sera from the biological duplicates were used (Author response image 2b). This antigenic cartography map is now added as Figure 2. Figure supplement 3 to the revised manuscript.

The antigenic cartography was constructed using averaged data from pairs of ferrets (a).

Similar analysis was performed on individual ferrets sera (b).

(3) Inconsistency in group labelling

A/Hatay/4990/2016 & A/New Caledonia/23/2016 are in phylogenetic group 1 in Figure 2 and phylogenetic group 1 in Figure 5 - figure supplement 1 panel a.

Our apologies: there was indeed a mistake in labeling of Figure 5. A new antigenic cartography was constructed and included in the revised manuscript. As a result Figure 5 - figure supplement has now become redundant and was removed from the manuscript.

A/Kansas/14/2017 is selected as a representative of antigenic group 2, when in Figure 2 it is labelled as AC1 (although Figure 2 - supplement 4 which the text is referring to shows data for A/Singapore/Infimh-16-0019/2016 as the representative of AC2). A/Kansas/14/2017 is coloured and labelled as AC2 in Figure 2 - supplement 5.

Thank you for pointing out this inconsistency. Kan17 clustered antigenically in group 1 based on the NAI values that were normalized relative to the serum with the maximal NAI value against the H6N2 virus that was tested. When using NAI titers that are normalization with the homologous ELISA titer, Kan17 is positioned in group 2. Likewise, antigenic cartography mapping positions Kan17 in group 2. Therefore, we conclude that A/Kansas/14/2017 NA is a representative of group 2.

The colouring is changed for Figure 3a at the bottom. A/Heilongjiang-Xiangyang/1134/2011 is coloured the same as AC4 viruses when it is AC1 in Figure 2. This lack of consistency makes the figures misleading.

We apologize for this mistake. The coloring in Figure 3a has been corrected.

(4) Data not presented, without explanation

The paper states that 44 sera and 27 H6N2 viruses were used (line 158). However, the results for the Kansas/14/2017 sera do not appear to be presented in any of the figures (e.g. Figure 2 phylogenetic tree, Figure 5 - figure supplement 1). It is not obvious why these data were not presented. The exclusion of this serum could affect the results as often the homologous titre is the highest and several heatmaps show the fold down from the highest titre.

Serum against A/Kansas/14/2017 was not prepared. For that reason, it is not included in the analysis. We agree that such homologous serum ideally should have been included and in the NAI assay would have resulted in a high if not the highest titer. However, we noticed that homologous sera did not always have the highest titers, especially in panels like ours were some antigens are closely related. The highest titer obtained against Kan17 H6N2 was from A/Bris/16 sera: 1/104, a titer that is in the range of other, homologous titers observed in the panel (Table S3). The Bris16 and Kan17 NAs have five amino acid differences. In summary, inclusion of Kan17 homologous sera would likely not impact the analysis and interpretation of the results because there are multiple highly cross-inhibiting heterologous serum samples against Kan17.

(5) The cMDS plot does not have sufficient quality assurance A cMDS plot is shown in Figure 5 - figure supplement 1, generated using classical MDS. The following support for the appropriateness of this visualisation is not given. a. Goodness of fit of the cMDS projection, including per point and per titre. b. Testing of the appropriate number of dimensions (the two sera from phylogenetic group 3 are clustered with phylogenetic group 2; additional dimensions might separate these groups). c. A measure of uncertainty in positioning, e.g. bootstrapping. d. A sensitivity analysis of the assumption about titres below the level of detection (i.e. that <20 = 10). Without this information, it is difficult to judge if the projection is reliable.

We agree with these comments. We have removed Figure 5 – figure supplement 1, and added new figure 2 – figure supplement 3 (antigenic cartography) instead.

(6) Choice of antigenic distance measure

The measure of antigenic distance used here is the average difference between titres for two sera. This is dependent on which viruses have been included in the analysis and will be biased by the unbalanced number of viruses in the different clusters (12, 8, 2, 5).

To verify the impact of the number of antigens on our analysis, the matrix of differences was generated with only 4 H6N2s representing at least one phylogenetic group (Per09, Sin16, Hel823 and Ind11) (Author response image 3a). This matrix is very similar to the one calculated based on all 27 antigens (Author response image 3b). The obtained matrix (Author response image 3a) was used in random forest to model antigenic distances and the result of prediction was plotted against real differences calculated based on the full data. The correlation coefficient (R2) of predicted vs observed values dropped from 0.81 to 0.71, suggesting that the number of antigens tested does not drastically affect the antigenic differences calculated based on serum values (Author response image 3e). Importantly, amino acid substitutions potentially associated with increased antigenic distances are similarly identified (Author response image 3c, d and f).

Matrix of differences was calculated using only 4 H6N2 antigens (a) or the full panel (b).

The matrixes from (c) 4 or (d) 27 antigens were used in random forest modeling to estimate the impact of amino acid changes, respectively. The rf modeling data generated from 4 H6N2 only was plotted and correlated with values calculated from the full panel of 27 H6N2s (e). The multi-way importance plot indicates in red that 7 out of the 10 most important substitutions were identified by the analysis using only 4 H6N2s (f).

Interestingly, when matrix of differences is calculated using only 4 H6N2s data but not including at least one representative of antigenic group 1 and 2, the correlation coefficient between the predicted values and values obtained from the full panel is dramatically impacted (R2 values drops from 0.81 to 0.5 and 0.57. It is important to note that most of the sera also belong to phylogenetic antigens from groups 1 and 2. As a consequence, poorer prediction of those antigens would more drastically impact the correlation. No drastic drop was observed when representative H6N2s from group 3 or 4 were excluded from the data (from 0.81 to 0.75 and 0.73, Author response image 4 c and d).

Random forest analysis was repeated using only 4 antigens, but excluding representatives of one of the phylogenetic groups (a) no group 1, (b) no group 2, (c) no group 3, and (d) no group 4.

We also used Euclidean distances as a measure of differences (Author response image 5). The predictive values obtained in rf have a slightly reduced R2 compared to the values obtained using average of differences.

In conclusion the unbalanced number of antigens used per group and metric of distance does not seem to impact per se our analysis.

Antigenic distances were calculated using Euclidian distances of sera to sera.

Those antigenic distances were used in rf for estimation of antigenic distance and importance of each amino acid substitution.

(7) Association analysis does not account for correlations

For each H6N2 virus and position, significance was calculated by comparing the titres between sera that did or did not have a change at that position. This does not take into account the correlations between positions. For haemagglutinin, it can be impossible to determine the true antigenic effects of such correlated substitutions with mutagenesis studies.

Most of the potential correlated effects cannot be addressed with the panel of N2s, except for combinations of substitution that are included in the panel, such as 245/247 with or without 468. Only mutagenesis studies would shed light on the epistatic effects. However, it is important to keep in mind that those individual substitutions in such kind of study likely do not reflect natural evolution of N2 (cfr. the importance of the NA charge balance (Wang et al., 2021: 10.7554/eLife.72516).

(8) Random forest method

25 features are used to classify 43 sera, which seems high (p/3 is typical for classification). By only considering mismatches, rather than the specific amino acid changes, some signals may be lost (for example, at a given position, one amino acid change might be neutral while another has a large antigenic effect). Features may be highly, or perfectly correlated, which will give them a lower reported importance and skew the results.

The number of features were optimized in the range from 5 to 80, with 25 being optimal (best R-value in predicted vs observed antigenic distances). Those features refer to the number of amino acid substitutions used in each tree. The number of trees was also optimized in the range of 100 to 2000.

In random forest the matrix of differences is made considering only position based and not the type of substitution in pairs of NA. Indeed, substitutions with distinct effects may skew results by indicating lower reported importance.

We have highlighted such potential bias in our discussion:

“Also, our modelling does not consider that substitution by other amino acids can have a distinct impact on the antigenic distance. As a consequence, predictions based on the model could underestimate or overestimate the importance of a particular amino acid residue substitution in some cases.”

Reviewer #2 (Public Review):

Summary:

The authors characterized the antigenicity of N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 using ferret and mice immune sera. Four antigenic groups were identified, which correlated with their respective phylogenic/ genetic groups. Among 102 amino acids differed by the 44 selected N2 proteins, the authors identified residues that differentiate the antigenicity of the four groups and constructed a machine-learning model that provides antigenic distance estimation. Three recent A(H3N2) vaccine strains were tested in the model but there was no experimental data to confirm the model prediction results.

Strengths:

This study used N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 and generated corresponding panels of ferret and mouse sera to react with the selected strains. The amount of experimental data for N2 antigenicity characterization is large enough for model building.

Weaknesses:

The main weakness is that the strategy of selecting 44 A(H3N2) viruses from 2009-2017 was not explained. It is not clear if they represent the overall genetic diversity of human A(H3N2) viruses circulating during this time. A comprehensive N2 phylogenetic tree of human A(H3N2) viruses from 2009-2017, with the selected 44 strains labeled in the tree, would be helpful to assess the representativeness of the strains included in the study.

The selection of antigens was performed using the method described by Bien and Tibshirani 2011 (doi: 10.1198/jasa.2011.tm10183). This method calculates MinMax distances to identify a central representative among distinct clusters.

To facilitate visualization of in a phylogenetic tree, only 180 representative N2 proteins from 2009-2017 were randomly selected (20 strains per year, unlabelled). Those 180 representatives and 44 readout panel strains (labelled) are shown in the phylogenetic tree below. Readout strains cover the major branches of the tree. The tree has been built using PhyML 3.0 using JTT substitution model and default parameters (Guindon S. et al, Systematic Biology 59(3):307-21, 2010) and visualized using ETE3 (Huerta-Cepas J. et al, Mol. Biol. Evol 33(6):1635-38, 2016).

The second weakness is the use of double-immune ferret sera (post-infection plus immunization with recombinant NA protein) or mouse sera (immunized twice with recombinant NA protein) to characterize the antigenicity of the selected A(H3N2) viruses. Conventionally, NA antigenicity is characterized using ferret sera after a single infection. Repeated influenza exposure in ferrets has been shown to enhance antibody binding affinity and may affect the cross-reactivity to heterologous strains (PMID: 29672713). The increased cross-reactivity is supported by the NAI titers shown in Table S3, as many of the double immune ferret sera showed the highest reactivity not against its own homologous virus but to heterologous strains. Although the authors used the post-infection ferret sera to characterize 5 viruses (Figure 2, Figure Supplement 4), the patterns did not correlate well. If the authors repeat the NA antigenic analysis using the post-infection ferret sera with lower cross-reactivity, will the authors be able to identify more antigenic groups instead of 4 groups?

This is a very valuable remark. In their paper, Kosikova et al. (CID 2018) report that repeated infection of ferrets with antigenically slightly different H3N2 viruses results in a broader anti-HA response, compared to a prime infection of an influenza naïve ferret, which results in a narrower anti-HA response. In our ferret immunizations the boost was performed with recombinant, enzymatically active NA that was homologous to the NA of the H1N2 virus that was used for the priming by infection. We determined the NAI responses in sera from ferrets after H1N2 infection against 5 different H6N2 viruses (Figure 2 – figure supplement 5). Compared to NAI responses in sera from H1N2 infected and subsequently NA protein boosted ferrets, the NAI titers obtained after a single infection were considerably lower. Although the normalized NAI titers of day 14 and day 42 sera correlated well, we cannot exclude a degree of broadening of the NAI response in the NA protein boost sera (Author response image 7). On the other hand, repeated influenza antigen exposure is the reality for the majority of people.

Correlation obtained on NAI data from ferrets at day 14 after infection vs data from day 42 after boost.

Another weakness is that the authors used the newly constructed model to predict the antigenic distance of three recent A(H3N2) viruses but there is no experimental data to validate their prediction (eg. if these viruses are indeed antigenically deviating from group 2 strains as concluded by the authors).

Indeed, there is no experimental data from A/Hong_Kong/45/2018, A/Tasmania/503/2020, or A/Darwin/9/2021. The generation of data to determine experimental values for A/Hong_Kong/45/2018, A/Tasmania/503/2020, or A/Darwin/9/2021 would require the generation of new reassortant viruses (H1N2s), recombinant protein and immunization of new ferrets. The ferrets sera would have to be analyzed against all 27 H6N2s, including duplicated control sera for normalization. The major point of the modeling was to evaluate if it is possible to predict the antigenic behavior based on amino acid substitutions.

As an exercise we have run the model again but this time excluding the Swe17 and HK17 antigens from the data set. Sequences of Sw17 or HK17 were then used to predict antigenic distances. The modeled versus experimental data are plotted in Author response image 8 and show a robust predictive outcome with R2 values of 0.94 and 0.91 for Sw17 and HK17, respectively.

Antigenic distances from Swe17 and HK17 calculated using the random forest algorithm that was constructed without experimental data from Swe17 and HK17.

The predicted distances were plotted side by side to the experimental distances in (a) and correlations are shown in (b).

Reviewer #3 (Public Review):

Summary:

This paper by Portela Catani et al examines the antigenic relationships (measured using monotypic ferret and mouse sera) across a panel of N2 genes from the past 14 years, along with the underlying sequence differences and phylogenetic relationships. This is a highly significant topic given the recent increased appreciation of the importance of NA as a vaccine target, and the relative lack of information about NA antigenic evolution compared with what is known about HA. Thus, these data will be of interest to those studying the antigenic evolution of influenza viruses. The methods used are generally quite sound, though there are a few addressable concerns that limit the confidence with which conclusions can be drawn from the data/analyses.

Strengths:

The significance of the work, and the (general) soundness of the methods.

Explicit comparison of results obtained with mouse and ferret sera.

Weaknesses:

Approach for assessing the influence of individual polymorphisms on antigenicity does not account for the potential effects of epistasis.

Indeed, possible epistatic effects or individual polymorphisms were not assessed, which is limited by the nature of the panel of N2s selected in the study. We now emphasize this in the discussion as follows:

“Also, our modelling does not consider that substitution by different amino acids can have distinct impact on antigenic distance. As a consequence, predictions based on the model could underestimate the importance of a particular amino acid residue substitution in some cases.”

Machine learning analyses were neither experimentally validated nor shown to be better than simple, phylogenetic-based inference.

This is a valid remark and indeed we have found a clear correlation between NAI cross reactivity and phylogenetic relatedness. However, besides achieving good prediction of the experimental data (as shown in Figure 5 and in FigureR7), machine Learning analysis has the potential to rank or indicate major antigenic divergences based on available sequences before it has consolidated as new clade. ML can also support the selection and design of broader reactive antigens.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Major corrections

No major corrections, beyond the issues I touched on in the public review, for which I give a little more detail below:

Point 2. If there's not a putative genetic basis for the unexpected clustering seen in the NAI, then reiterating a small subset of the data would show the reliability of the experimental methods and substantiate this unexpected finding.

We thank the reviewer for this pertinent point and suggestion. We have modified our analysis by reiterating individual ferret data normalized with the homologous ELISA titers. This reiteration is shown in figure R1b. In this case both Kan17 and Wis15 are switched to antigenic group 2. The profile of sera inhibition against those 2 strains that shift from antigenic cluster 1 to 2, is clearly an intermediate between profiles observed in those 2 groups. Considering that antigenic evolution occurs gradually, it is not unexpected that those intermediate profiles would swing from one side to another when pushed to forced discrimination. Antigenic cartography mapping, as in Smith et al. (2004), also indicated that those H6N2s are located closer to G1 than overall antigens from G2. Raw data distribution (max and min EC50) also do not indicate potential bias in analysis.

Point 5. If you want to use antigenic cartography (Smith et al 2004), there is the R CRAN package (https://CRAN.R-project.org/package=Racmacs) which can handle threshold titres (like <20) and has functions for the diagnostic tools I describe, in order to quality assure the resulting plot. It does use a different antigenic distance metric than the paper currently uses, so you might not want to take that route.

Thank you for this suggestion. We have performed antigenic cartography using the methodology described by Smith et al made accessible by Sam Wilks. The outcome of this analysis has been added to the manuscript as Figure 2 – Figure supplement 3.

Point 6. More robust measures of antigenic distance take into account the homologous titre, homologous and heterologous titres (Archetti & Horsfall, 1950) or use the highest observed titre for a serum (Smith et al 2004). A limitation of the first two is that the antigenic distance can only be calculated when you have the homologous titre, which will limit you as you only have this for 26/43 sera. They may give similar results to your average antigenic distance, in which case your analysis still stands. Calculating antigenic distance using the homologous or maximum titre only gives the antigenic distance between the antigen and the serum. If you want the distance between all the sera, then further analysis is required (making an antigenic map and outputting the serum-serum distances, see the point above).

We thank the reviewer for these suggestions. A complete set of 43 H6N2 viruses that matches all 43 sera would have been ideal. This would require the generation of 17 additional H6N2 viruses and their testing in ELLA, a significant amount of work in terms of time and resources. Instead, we have generated an antigenic map of the 27 antigens and homologous sera (cfr. our response to point 5 above). Despite different methods the outcome showing 4 major antigenic groups is consistent.

Minor corrections

Table S1

A/New_Castle/67/2016 should be A/Newcastle/67/2016

A/Gambia/2012 is not the full virus name

Corrected.

Table S3 has multiple values of exactly 10.0. I think these should be <20 as they are below the threshold of detection for the assay.

All the values lower than 20 in Table S3 were replaced by “< 20”.

Line 376: A/Sidney/5/1997 should be A/Sydney/5/1997

Corrected.

Line 338: "25 randomly sampled data" is a bit vague, "25 randomly sampled features" would be better

Corrected.

Include RMSE of the random forest model.

RMSE=19.6 RMSE/mean = 0.207 is now mentioned in the manuscript.

Figure 5 - supplement 1: These plots are difficult to interpret as the aspect ratio is not 1:1, and panels a & b are difficult to compare as they have not been aligned (using a Procrustes analysis). It would be neater if they were labelled with short names.

We have generated an antigenic cartography map instead. As a consequence, the MDS has become redundant and Figure 5 – supplement 1 was removed.

Line 562: 98 variable residues, where it is 102 elsewhere in the text.

There are 4 mutations near the end of the NA stalk domain, which are not resolved in the N2 structure. Therefore, amino acid distances to these residues cannot be calculated.

No data availability statement. Some of the raw data is available in Table S3 and there is no link to the code.

The data and code used for generation of rf modelling was uploaded to Github and made available. The following statement has been added to the manuscript: “The data and code used for the generation of the rf model is available at https://github.com/SaelensLAB/RF..”

Reviewer #2 (Recommendations For The Authors):

(1) More than 42,000 NA sequences are available for the mentioned period on GISAID, it is therefore important to understand the selection criteria for the 44 strains and if these strains represent the overall genetic diversity of N2 of human A(H3N2) viruses. To demonstrate the representativeness of the 44 selected strains, please construct a representative N2 phylogenetic tree for human A(H3N2) viruses circulated in 2009-2017 and label the 44 selected strains on the tree.

The selection of antigens was performed using the method described by Bien and Tibshirani 2011 (doi: 10.1198/jasa.2011.tm10183). This method uses MinMax distances to identify a central representative among distinct clusters.

To facilitate visualization tree only of 180 representative N2 proteins from 2009-2017 were randomly selected (20 strains per year, unlabelled). Those 180 representatives and 44 readout panel strains (labelled) are shown in the phylogenetic tree below. Readout strains cover the major branches of the tree. The tree has been built using PhyML 3.0 using JTT substitution model and default parameters (Guindon S. et al, Systematic Biology 59(3):307-21, 2010) and visualized using ETE3 (Huerta-Cepas J. et al, Mol. Biol. Evol 33(6):1635-38, 2016).

(2) Double immune ferret sera may increase antibody binding affinity and cross-reactivity against heterologous strains. Using single-infection ferret sera may yield different antigenic grouping results (eg. may identify more antigenic groups). Can the authors repeat the NA antigenic grouping using single-infection ferret sera? Although data from a subset of 5 strains was presented (Figure 2, Figure Supplement 4), the information was not sufficient to support if the use of single-infection or double immune ferret sera will yield similar antigenic grouping results.

In our ferret immunizations the boost was performed with recombinant, enzymatically active NA that was homologous to the NA of the H1N2 virus that was used for the priming by infection. We determined the NAI responses in sera from ferrets after H1N2 infection against 5 different H6N2 viruses (Figure 2 – figure supplement 5). Compared to NAI responses in sera from H1N2 infected and subsequently NA protein boosted ferrets, the NAI titers obtained after a single infection were considerably lower. Although the normalized NAI titers of day 14 and day 42 sera correlated well, we cannot exclude a degree of broadening of the NAI response in the NA protein boost sera (Figure R6). On the other hand, repeated influenza antigen exposure is the reality for the majority of people.

(3) NA antigenicity data is presented in heat maps and the authors would often describe the heat map patterns matches without further explanations. Line 234-235, the heat map of mouse sera (Figure 2. Figure supplement 5) was described to match the results of ferret sera (Figure 2), but this tends to be subjective. A correlation analysis of 7 selected antigens showed a positive correlation, what about the other 37 antigens?

The interpretation of heatmaps is indeed very subjective, for this reason the correlation of the 7 selected antigens was also provided. The other 37 antigens were not tested. Considering the results using post boost sera, a simulation of using random forest modeling indicate that the data from one antigen of each antigenic group is sufficient to achieve a reliable predictive output (R2=0.71) (Figure R3 of this rebuttal).

(4) Can the authors explain in more detail how data in Figure 4a was generated? According to the authors, residues close to the catalytic pocket are more likely to impact NAI. Can the authors explain how they define if a residue is close to the catalytic pocket?

The correlation of distances of amino acid residues with significance values is explained as follows. Consider 7 distinct elements that are distributed horizontally as shown by the squares in the figure below (Author response image 10a). The elements highlighted in yellow have a numerical propriety (in case of N2 neuraminidase this was the significance values obtained in the association study). Taking P1 as reference we can calculate the distance (red arrows) between P1 and P2, P4 and P7, those distances can them be correlated to intrinsic values of P2, P4 and P7, which enables the calculation of the correlation coefficient Tau. This same process is repeated for each position (or each amino acid), as a consequence every position will have a correlation coefficient calculated (Author response image 8b). This correlation coefficient can be represented as a heat map at the surface of N2.

The 2D scheme represents the strategy used to calculate the correlation (i.

e. the Tau values) between distances and p-values. Tau values can then be presented in a heat map.

(5) Can the authors provide experimental data using the three recent A(H3N2) viruses as antigens and perform NAI assay to confirm if they are antigenic all deviating from group 2 viruses?

The generation of data to determine experimental values for A/Hong_Kong/45/2018, A/Tasmania/503/2020, or A/Darwin/9/2021 would require the generation of new reassortant viruses (H1N2s), recombinant protein and immunization of new ferrets. The ferrets sera would have to be analyzed against all 27 H6N2s, including duplicated control sera for normalization. The major point of the modeling was to evaluate if it is possible to predict the antigenic behavior based on amino acid substitutions.

As an exercise we have run the model again but this time excluding the Swe17 and HK17 antigens from the data set. Sequences of Sw17 or HK17 were then used to predict antigenic distances. The modeled versus experimental data are plotted in Author response image 7 and show a robust predictive outcome with R2 values of 0.94 and 0.91 for Sw17 and HK17, respectively.

(6) According to Ge et al. 2022 (PMID: 35387078), N2 NA's before 2014 (2007-2013) showed a 329-N-glycosylation and E344, and they were subsequently replaced by H3N2 viruses with E344K and 329 non-glycosylation changing the NI reactivity in ferret antisera towards later strains. Were these residues also predicted to be important to N2 antigenicity from your machine-learning method?

Three of the N2 NAs used in our panel, A/Victoria/361/2011, A/Hong_Kong/3089/2017, and A/Tennessee/18/2017, lack this N-glycosylation motif. The E344K substitution is present in another 3 NAs, derived from A/Nagano/2153/2017, A/Minnesota/11/2010, and A/Indiana/08/2011. The importance of those mutations is among the lowest ones predicted in our modeling. However, the differences in NAI reported by Ge et al. are low (not even twofold). The experimental variability in our study potentially limits the identification of substitutions with a subtle impact NAI. We have added the following to the discussion in our revised manuscript:

“It has been reported that an N-glycosylation site at position 329 combined with E344 in NA from human H3N2 viruses from 2007 to 2013 was gradually lost in later H3N2 viruses (Ge et al., 2022). This loss of an N-glycosylation site at position 329 combined with an E344K substitution was associated with a change in NAI reactivity in ferret sera. Three N2 NAs in our panel, derived from A/Victoria/361/2011, A/Hong_Kong/3089/2017, and A/Tennessee/18/2017, lack this N-glycosylation motif. The E344K substitution is present in three other NAs, derived from A/Nagano/2153/2017, A/Minnesota/11/2010, and A/Indiana/08/2011. The importance of those mutations is among the lowest ones predicted by our modeling. However, the differences in NAI reported by Ge et al. are very modest (lower than twofold). The experimental variability in our study potentially limits the identification of substitutions with a subtle impact NAI.”

Reviewer #3 (Recommendations For The Authors):

Specific suggestions:

Line 132: Did the authors confirm the absence of compensatory mutations due to a heterologous H6 background that could potentially confound downstream NAI results?

All NAs genes of the rescued H6N2 viruses were fully sequenced and were found to be identical to the expected NA sequences, with the only exception being the A/Tasmania/1018/2015 were a mixed population of wt and M467I was found. This substitution is located at the surface and at the top of the NA head domain, and thus could potentially impact NA antigenicity. However, A/Tasmania/1018/2015 H6N2s had a similar inhibition profile as other H6N2s in phylogenetic and antigenic group 1. This indicates that, at least in this mixed population, antigenicity was not drastically affected by the M467I substitution.

Line 96: how do these data rule out variation in the fraction of properly folded protein across NAs? They certainly show that properly folded NA protein is present, but not whether amounts vary between the different NAs.

SEC-MALS (size exclusion chromatography-Multiangle light scattering) data and enzymatic activity were considered as a proxy for correctly folded NA. Although the specific activity of the recombinant N2 NAs is expressed per mass unit (microgram), we cannot exclude that the fraction of properly folded protein across the different recombinant NAs may vary.

Lines 262-269: this analysis approach (based on my reading) seems to consider each polymorphism in isolation and thus does not seem well suited for accounting for epistatic interactions within the NA. For example, the effect of a substitution on NAI may be contingent upon other alleles within NA that are not cleanly segregated between the two serum comparator groups. Can the authors address the potential of epistasis within NA to confound the results shown in Figure 3?

Unfortunately, epistatic interactions cannot be solved using the panel of N2 selected for the study. This limitation is mentioned in our discussion:

“It is important to highlight that co-occurring substitutions in our panel (the ones present in the main branches of the phylogenetic tree) cannot be individually assessed by association analysis or the random forest model. The individual weight of those mutation on NA drift thus remains to be experimentally demonstrated.”

Line 331: is there a way to visualize and/or quantify how these two plots (F5 supplement 1a/b) reflect each other or not? Without this, it is hard to ascertain how they relate to each other.

We have generated an antigenic cartography map instead. As a consequence, the MDS has become redundant and Figure 5 – supplement 1 was removed.

Figure 4B structural images are not well labelled.

The active site in 1 of the protomers is now indicated with an arrow in the top and side views of the NA tetramer.

Lines 339-359: the ML predictions are just predictions and kind of meaningless without experimental validation of the predicted antigenic differences between recent NAs. This section would also be strengthened by an assessment of whether the ML approach obtains more accurate results than simply using phylogeny to predict antigenic relationships.

Indeed, there is no experimental data from A/Hong_Kong/45/2018, A/Tasmania/503/2020, or A/Darwin/9/2021. The generation of data to determine experimental values for A/Hong_Kong/45/2018, A/Tasmania/503/2020, or A/Darwin/9/2021 would require the generation of new reassortant viruses (H1N2s), recombinant protein and immunization of new ferrets. The ferrets sera would have to be analyzed against all 27 H6N2s, including duplicated control sera for normalization. The major point of the modeling was to evaluate if it is possible to predict the antigenic behavior based on amino acid substitutions.

As an exercise we have run the model again but this time excluding the Swe17 and HK17 antigens from the data set. Sequences of Sw17 or HK17 were then used to predict antigenic distances. The modeled versus experimental data are plotted in figure R7 and show a robust predictive outcome with R2 values of 0.94 and 0.91 for Sw17 and HK17, respectively. A major advantage of antigenic modeling is the potential to rank or indicate major antigenic divergences based on available sequences before it has consolidated as new clade. The support in selecting or designing broader reactive antigens is another advantage of machine learning analysis.

Lines 416-421: appreciate the direct comparison of results obtained from ferrets versus mice.

We thank the reviewer for expressing this appreciation.

Version 2

Abstract

Abstract

Human H3N2 influenza viruses are subject to rapid antigenic evolution which translates into frequent updates of the composition of seasonal influenza vaccines. Despite these updates, the effectiveness of influenza vaccines against H3N2-associated disease is suboptimal. Seasonal influenza vaccines primarily induce hemagglutinin-specific antibody responses. However, antibodies directed against influenza neuraminidase (NA) also contribute to protection. Here, we analyzed the antigenic diversity of a panel of N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. The antigenic breadth of these NAs was determined based on the NA inhibition (NAI) of a broad panel of ferret and mouse immune sera that were raised by infection and recombinant N2 NA immunization. This assessment allowed us to distinguish at least 4 antigenic groups in the N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. Computational analysis further revealed that the amino acid residues in N2 NA that have a major impact on susceptibility to NAI by immune sera are in proximity of the catalytic site. Finally, a machine learning method was developed that allowed to accurately predict the impact of mutations that are present in our N2 NA panel on NAI. These findings have important implications for the renewed interest to develop improved influenza vaccines based on the inclusion of a protective NA antigen formulation.

Introduction

Neuraminidase (NA) and hemagglutinin (HA) constitute the two major antigenic proteins at the surface of influenza A and B virions (Uyeki et al., 2022). While HA mediates binding to sialic acids and membrane fusion, NA catalyses the removal of sialic acid residues from sialylated glycoproteins and glycolipids. This complementary role of HA and NA enables influenza virions to penetrate the mucus barrier that lines the respiratory epithelium and newly formed virions to be released from the infected cell (de Vries et al., 2020; Matrosovich et al., 2004; Palese et al., 1974). Immunity against influenza is associated with the presence of anti-HA antibodies in serum, as determined by a hemagglutination inhibition assay. In line with this, the composition of licensed influenza vaccines is primarily focused on the antigenic characteristics of HA as the protective vaccine antigen (Cox, 2013). Multiple studies, however, have underscored the protective potential of anti-NA immune responses as determined by NA inhibition (NAI) assays (Couch et al., 2013; Maier et al., 2019; Memoli et al., 2016; Monto and Kendal, 1973; Monto et al., 2015; Ng et al., 2019; Weiss et al., 2019). There is also evidence that pre-existing anti-N2 immunity induced by prior exposure to H2N2 viruses that circulated in the population from 1957 to 1968, tempered the impact of the H3N2 pandemic virus that emerged in 1968 (Eickhoff and Meiklejohn, 1969; Jester et al., 2020; Monto and Kendal, 1973; Viboud et al., 2005). Further, the pharmacological inhibition of NA activity is a well-established therapeutic strategy to try to reduce influenza severity and duration (Leneva et al., 2000; Yen et al., 2005). In addition, antibodies with NAI activity can protect against influenza virus infection in pre-clinical models (Chen et al., 2018; Deroo et al., 1996; Rosu et al., 2022; Wan et al., 2018; Wohlbold et al., 2017).

Since its appearance in 1968, human H3N2 viruses have been evolving faster than the co-circulating H1N1 and influenza B viruses (Bedford et al., 2014; Potter et al., 2019). The overall Influenza virus vaccine effectiveness varies between seasons and, despite frequent updates of the H3N2 HA component, the effectiveness of influenza vaccines against H3N2-associated disease ranges between 5 and 39%, which is lower than the effectiveness against H1N1 and influenza B (Belongia et al., 2016). In this context the use of NA as a complementary vaccine antigen might improve influenza vaccine effectiveness, for example by compensating for HA-based antigenic mismatches in circulating influenza virus strains (Krammer et al., 2018).

Phylogenetic analyses of HA and NA genes of human H3N2 viruses revealed the asynchronous evolution of the 2 proteins. In addition, the evolutionary rate as determined by the accumulation of mutations over time, is estimated to be 1.5 times lower for NA than for HA (Westgeest et al., 2012). Despite the strong evidence that immune responses directed against NA can contribute to protection against influenza, a comprehensive insight in the antigenic breadth and evolution of the H3N2 NA antigen is lacking (Nachbagauer et al., 2017; Rajendran et al., 2017; Smith et al., 2004; Westgeest et al., 2012).

Here, we determined the antigenic landscape of N2 NAs derived from human H3N2 strains that were isolated between 2009 and 2017. This analysis allowed the identification of at least 4 N2 NA antigenic groups and to estimate the weight of amino acid changes on their contribution to NA antigenic drift.

Results

Generation of H3N2-derived NA immune sera in ferrets and mice

Our aim was to determine the antigenic relationship between NAs derived from human H3N2 viruses that circulated during nearly one decade. We first selected 43 H3N2 neuraminidase protein sequences obtained via GISAID, derived from viruses that were isolated between 2009 and 2017 (Elbe and Buckland-Merrett, 2017). Hierarchical clustering with minimax linkage was employed to ensure the set was representative of overall protein sequence diversity (Bien and Tibshirani, 2011). Next, 43 recombinant H1N2 viruses were generated that carried the internal and HA genes from PR8 virus and the NA genes of the selected H3N2 viruses. In addition, the corresponding tetrabrachion-stabilized recombinant N2 NA proteins were produced in Chinese hamster ovary (CHO) cells and purified, and are referred to as NACHOs (Catani et al., 2022). All purified NACHOs were tetrameric based on SEC-MALS analysis. The enzymatic activity of the NACHOs was determined based on the hydrolysis of the fluorogenic small molecule substrate 4-methylumbelliferyl-N-acetylneuraminic acid (MUNANA). The NACHOs were active with a specific activity ranging from 0.27 to 27.48 nmole/min/µg, indicating correct folding and assembly of the NA head domains (Table S1).

Immune sera against all 43 H3N2-derived NAs were generated in pairs of ferrets by a primary infection with the respective H1N2 reassortant viruses. This prime by infection was used to avoid the use of an adjuvant upon subsequent boost with the corresponding NACHOs. On day 28 after primary infection, all animals were boosted by intramuscular injection of homologous NACHO, except for the ferrets that had been primed by infection with A/Stockholm/15/2014 and A/Estonia/91621/2015, which were re-inoculated with the same respective H1N2 reassortant viruses. These two pairs of ferrets were not boosted with homologous NACHO because the corresponding NACHOs were not available at the time of the experiment. Blood samples were taken from the ferrets on days 14 and 42 after (primary) infection (Figure 1a). Priming by H1N2 infection resulted in seroconversion against the respective immobilized homologous NACHOs (Figure 1b). Subsequent immunization with recombinant NACHOs boosted the NACHO-specific IgG titers in day 42 serum samples. In contrast, serum samples from the ferrets that had been re-inoculated with H1N2 reassortant carrying NA derived from A/Stockholm/15/2014 or A/Estonia/91621/2015, showed no further increase in titers, presumably because prior exposure with the same H1N2 virus had induced a sterilizing immune response (Figure 1b).

Generation of anti-N2 sera panel from ferrets and mice.

(a) 43 Ferret immunization and blood sampling scheme. The second immunization on day 28 was performed with homologous NACHO except for the ferrets that had been primed with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014, which were infected again on day 28 post prime with the same H1N2 viruses. (b) Homologous NACHO ELISA titer of ferret sera sampled on day 14 after inoculation with H1N2 virus and 2 weeks after the boost with recombinant NA or by re-inoculation with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014 (indicated by arrows on the X-axis). (c) 43 BALB/c mouse immunization and blood sampling scheme. (d) Homologous NACHO serum IgG titers in pooled sera from mice that were immunized twice with recombinant NA. Mouse and ferrets representations were created with BioRender.com (2023).

We also generated NA immune sera in BALB/c mice. Groups of 5 mice were immunized two times, with a two-week interval, by intramuscular injection with 1 µg of the respective 43 NACHOs adjuvanted with AF03. Blood was collected from the immunized mice and serum prepared on day 28 after the prime immunization (Figure 1c). NACHO-specific IgG titers in the mouse immune sera, pooled from each group, ranged from 1:8100 to 1:72900 (Figure 1d).

The antigenic profile of N2 neuraminidase

Next, NAI titers were determined in the ferret sera using a panel of H6N2 reassortant viruses in an enzyme-linked lectin assay (ELLA), avoiding interference by anti-H1 HA antibodies. We generated 27 H6N2 influenza viruses with H6 HA derived from A/mallard/Sweden/2002 and NA derived from a subset of the 43 H3N2 viruses, thereby retaining NA sequence diversity and year of H3N2 virus isolation (Table S2).

The 43 NAs used as antigens and 27 NAs as source of NA, could be classified into 4 groups when ordered according to the phylogenetic relatedness of their NA sequence (Figure 2). The conversion rate of MUNANA into 4-methylumbelliferyl and acetylneuraminic acid by these H6N2 viruses revealed a KM that ranged from 20.7 to 144.8 µM as determined by the Michaelis-Menten equation (Figure 2 – figure supplement 1a and Table S2). Interestingly, the KM differs between the distinct phylogenetic groups (Figure 2 – figure supplement 1a). There was a negative correlation between the KM values and the dilution of the H6N2 viruses required to set the 70% maximal NA activity in ELLA (Figure 2 – figure supplement 1b).

Breadth of N2 NA inhibition.

43 Phylogenetic tree based on the N2 NA head domain amino acid sequences and heat-map representing the average of normalized neuraminidase inhibition per H6N2 [log2 (max (NAI/ ELISA homologous titer)/(NAI/ homologous ELISA titer)] determined in ferret sera after the boost (listed vertically). The red-to-blue scale indicates high-to-low NAI observed in ELLA against the H6N2 reassortants (listed at the bottom). UPGMA clustering of H6N2s inhibition profiles are shown on top of the heat map and coloured according to the phylogenetic groups.

Using this panel of 27 H6N2s, we determined the NAI titers in the sera obtained from the 43 pairs of ferrets 2 weeks after the boost. ELLAs were performed in duplicate for all serum samples against one H6N2 target at the same time. The reciprocal NAI titers of the ferret sera paired with H6N2 viruses are shown in Table S3. In general, the NAI titers of the ferret sera were the highest against homologous and phylogenetically closely related H6N2 viruses. In Figure 2, all the tested ferret sera are represented in the rows and the 27 H6N2 viruses against which the sera were evaluated in ELLA are presented in the columns. The serum samples were ordered according to the phylogenetic relatedness of the N2 NA head domain amino acid sequences, showing that these cluster in 4 groups (G1-G4). The ELLA-based IC50 NAI titers were normalized by the homologous ELISA titers and then plotted relative to the maximum NAI titer obtained against the respective H6N2 target viruses. The heatmap thus represents, for each H6N2 virus, the normalized ELLA IC50 titer expressed as log2 of the fold-difference relative to the serum with the highest IC50 value against that H6N2 virus (dark red squares indicate high and blue squares low IC50 values) (Figure 2).

Next, these normalized NAI values were clustered against each of the H6N2s using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA), which is a distance-based method. This clustering allowed to discern at least 4 major antigenic clusters (AC1 to −4) that are depicted at the top of the heat map (Figure 2). It is interesting to note that the H6N2 viruses with NA antigens derived from A/Wisconsin/16/15 and A/Alaska/251/15 do not segregate into AC2, as would be predicted based on the phylogenetic relatedness, but cluster in AC1 based on the UPGMA. However, the overall concordance of the antigenicity with the phylogenetic relatedness, based on the amino acid sequences of the NA head domains, was confirmed by hierarchical clustering with bootstrapping and k-means (Figure 2 – figure supplements 2a and 2b).

In addition, antigenic cartography analysis, based on the method described by Smith et al. (Smith et al., 2004), resulted in a antigenic map of the N2 NAs that was in line with the 4 proposed antigenic clusters (Figure 2 – figure supplement 3).

In line with this, a positive correlation was observed between the pairwise phylogenetic distances and the normalized NAI IC50 values for each of the tested H6N2s (Figure 2 – figure supplement 4).

We further tested the NAI activity of ferrets’ sera obtained 2 weeks after the prime infection against at least one H6N2 virus from each antigenic cluster. These H6N2 viruses were selected based on their antigenic relatedness as deduced from the ferret immune sera analysis (Figure 2) with 2 representatives from group 1 (Tex12, and Per09) and 1 each from group 2 (Kan17), group 3 (Hel823), and group 4 (Ind11). The normalized data confirmed the antigenic clustering observed in sera after boost (Figure 2 – figure supplement 5).

The NAI titers were also determined in the sera obtained from the 43 groups of NACHO-immunized mice against 7 of the H6N2 viruses (Figure 2 – figure supplement 6a). These H6N2 viruses were selected as described above with the addition of HK14 and Swi13 H6N2s, both belonging to antigenic cluster 1. Serum NAI IC50 values against a given H6N2 virus were normalized against the IC50 of the mouse serum that had the highest NAI titers against the tested H6N2 virus. UPGMA-based clustering showed that the mouse NACHO immune sera displayed a similar recognition pattern as the ferret immune sera that had been generated by H1N2 priming followed by a NACHO boost (Figure 2). For example, HK14 H6N2 NA activity is strongly inhibited by sera from group 1, is poorly inhibited by most of the sera from group 2 and is hardly inhibited by group 3 and 4 sera from both ferrets and mice.

Ind11 H6N2 NA activity, on the other hand, was inhibited by group 1 and 4, and hardly by group 2 and 3 immune sera. Overall, the normalized NAI IC50 values of the ferret and mouse immune sera correlated very well with coefficients values ranging from 0.33 to 0.78 (Figure 2 – figure supplement 6b).

Association analysis reveals amino acid residue substitutions that strongly impact NAI

NA of human influenza viruses is subject to antigenic drift, predominantly driven by the NAI activity that is detectable in the host serum (Air et al., 1985; Yasuhara et al., 2019). There are a total of 102 amino acid differences between the head domains of the NAs that were used to generate the 43 ferret immune sera (Figure 3 - Figure supplement 1). These 102 variable amino acids in the N2 head domain are predominantly exposed on the NA surface (Figure 3 – figure supplement 2).

Association analysis reveals impact of amino acid mutations on NAI.

(a) Association study performed for each H6N2 virus (listed at the bottom) to determine the probability of each variable amino acid (listed on the right) present in the panel of 43 NAs to impact the NAI. The red-to-blue scale represents significance values (S=-log10P) associated with each amino acid substitution. The dark red in the color scale represent all values that are equal or higher than 3 (corresponding to p-values ≤ 0.001). The colour shades on the bottom side indicate the distinct phylogenetic groups. (b) The significant values (p<0.05) were represented on the surface of N2 tetramer (PDB accession number 4H53) for one representative strain of each phylogenetic group with a yellow to red scale of significance (-log10P).

To try to identify which of these amino acid substitutions have a significant impact on NAI, we performed an association analysis for each of the tested H6N2 viruses. For this analysis, the NAI IC50 value was compared between groups of N2 sera according to the presence or absence of a substitution of any of the 102 amino acid position that are different between the panel of 43 tested N2s against which sera had been raised in ferrets. As a result, a p-value is obtained for each position that reflects its possible impact on NAI (Figure 3a). The significance values higher than 1.3 (p<0.05) of representative strains of the 4 antigenic groups were represented on the surface of N2 (Figure 3b). No obvious antigenic patches, however, could be distinguished on the N2 NA surface.

A similar association analysis was performed using the NAI values obtained with the mouse NACHO immune sera. The calculated p-values obtained from the mouse and ferret NAI data were in line with each other (Figure 3 – figure supplement 3).

Residues close to the catalytic pocket are more likely to impact NAI

We next calculated the distance between the Cα atom of each amino acid present in the N2 structure (PDB: 6BR6, from A/Perth/16/2009 NA) to each Cα of the variable amino acids present in the panel. We then determined the correlation of those distances with the p-values that were previously determined in the association analysis. The resulting correlation coefficients (Tau values) indicate a similar pattern among all 27 tested H6N2s (Figure 4a). When visualized on the N2 surface structure, the Tau values revealed a geometrical distribution with the negative correlations being in closer proximity to the catalytic pocket (Figure 4b). Similar data was obtained when mice sera were analyzed this way (Figure 4 – figure supplement 1).

Residues close to the catalytic pocket are more likely to impact NAI.

(a) Significance values (102 variable amino acids) obtained from the association study were correlated to the distance between every amino acid and each of the variable amino acids in the NA structure. The red-to-blue scale represents the Tau values obtained in the Kendall correlation. Amino acids 83 to 469 are shown from top to bottom in the heatmap. The H6N2s used in the NAI panel are listed at the bottom. The grey column indicates the residues that are part of the catalytic pocket (CP) that directly interact with the substrate (R118, D151, R152, R224, E276, R292, R371, and Y406) and the residues that play a key role in stabilizing the catalytic site (E119, R156, W178, S179, D198, I222, E227, H274, E277, N294, and E425). (b) Tau values obtained from Tex12 H6N2 correlation are represented on the surface of the N2 structure of A/RI/5+/1957 (PBD 4H53), the arrow indicates the catalytic pocket in one NA protomer.

Prediction of the N2 antigenicity based on a machine learning method

The antigenic drift of human H3 HA has been extensively studied and modelled (Meyer and Wilke, 2015; Smith et al., 2004; Westgeest et al., 2012; Wu et al., 2020). To model the NA drift observed in our panel of N2 NAs, we first calculated the antigenic distances between the distinct NA sera, using a similar approach as has been used to visualize the HA antigenic map (Cai et al., 2012). Then, using the antigenic distances and a random forest method (RF), a model was built based on amino acid sequence information. A similar model has been used by Yao et al. to predict the antigenicity of H3 HA (Yao et al., 2017). Here, the optimized model was built using a combination of 25 randomly sampled features in 1500 decision trees. The resulting model generates predictive antigenic distances that strongly correlate with the observed values (Figure 5a), with an RMSE/mean=0.207). The RF also enables scoring of amino acid residues according to their importance in the decision trees. One measure of importance can be defined as how early a change in a position is placed in the generated decision trees (mean minimal depth). Another measure of importance is how much the standard error mean increases when a variable is permuted. Finally, the prevalence of a variable amino acid in nodes of the decision tree can also indicate its importance. Those importance scores were represented in a multi-way plot (Figure 5b). The importance scores highlighted similar positions as previously determined in the association study. Finally, we used NA sequences of recent seasonal influenza vaccine recommended H3N2 strains to predict their antigenic distance relative to our panel of N2s, including 3 recent N2 NA components (from A/HongKong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021) that were not present in our data set. The predicted antigenic distances were plotted in the heatmap in a red to blue scale, indicating high to low antigenic distance, respectively (Figure 5c). The RF clearly reproduced the drift observed in AC2 versus AC1. The modelling also indicates that the antigenic distance of the 3 recent N2s to their corresponding group 2 NAs is increasing (represented by the dark blue turning light blue, from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G2) whereas their distance to group 4 NAs is decreasing (dark red becoming yellow from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G4).

Prediction of N2 antigenicity based on machine learning.

(a) Predicted vs observed normalized NAIs are shown. (b) Multi-way importance plot indicating the increase in mean standard error, the mean depth, and number of nodes. The top 10 most important amino acid changes are highlighted in blue and annotated with their position in the NA sequence. (c) 43 The heatmap indicates the predicted antigenic distances of N2s from recent H3N2 isolates (at the bottom of the heatmap) relative to the N2s that were used in the NAI breadth panel (listed vertically). The antigenic distances were calculated based on NA protein sequence.

Discussion

The sequence diversity of N2 proteins derived from the 43 selected human H3N2 viruses isolated between 2009 and 2017 allows to classify them in 4 major phylogenetic groups. Group 1 contains 22 N2 NAs, 4 of which were derived from vaccine recommended H3N2 strains: A/Perth/16/2009, A/Texas/50/2012, A/Switzerland/9715293/2013, and A/Hong Kong/4801/2014. Phylogenetic group 2 comprises 14 N2 NAs, one of which is derived from a vaccine H3N2 recommended strain:

A/Singapore/INFIMH-16-0019/2016. The third phylogenetic group contains two NAs that are derived from H3N2 strains that had been isolated in Helsinki in 2013. Remarkably, the NA sequence of these two strains is closely related to that of H3N2s that circulated more than 15 years earlier (e.g. 99% identity with NA of A/Sydney/5/1997). The fourth phylogenetic group is composed of NAs derived from zoonotic cases with swine H3N2 strains isolated between 2012 and 2017. Swine-to-human transmission of H3N2 viruses have occurred multiple times in the USA in 2012 and continue to be detected at low frequency until to date (CDC, 2023; Elbe and Buckland-Merrett, 2017).

We assessed the antigenic diversity of the NA panel by generating 27 recombinant H6N2 viruses and 43 ferret NA immune sera produced by H1N2 priming followed by boosting with homologous recombinant NA (NACHO; n = 41) or by reinfection with H1N2 (A/Estonia/91621/2015 and A/Stockholm/15/2014). Most of the ferrets had detectable serum anti-N2 antibodies by day 14 after infection with H1N2. Anti-NA antibody titers increased after a boost with non-adjuvanted homologous recombinant CHO-produced soluble tetrameric NA suggesting that the use of recombinant NA may boost naturally acquired anti-NA antibody responses in the human population.

The NAI activity of the ferret sera against 27 tested H6N2s reassortants revealed a pattern of inhibition that largely coincides with the phylogenetic distances based on the NA head domain sequences. In the tested panel of H6N2s, at least 4 major antigenic groups were identified. We noted that H6N2 with Kan17 NA had a broader profile of inhibition than other H6N2 viruses from group 2, being also inhibited by serum raised against NA from phylogenetic groups 3 and, to a lesser extent, group 4. Additionally, the H6N2s Wis15 and Ala15 have an antigenic behaviour that is more closely related to antigenic cluster 1, despite belonging to group 2 according to their phylogenetic similarities. Based on the NAI panel, an association analysis was performed to try to identify amino acid changes that affect NAI. It is interesting to note that the amino acid substitutions S245N and S247T, combined with substitution P468H, were identified in this association analysis. Those substitutions were described as major contributors of the NA antigenic drift that occurred in H3N2 in 2014 (Wan et al., 2019). NAs from A/Wisconsin/16/2015 and A/Alaska/251/2015 also have the S245N and S247T substitutions and phylogenetically belong to group 2 NAs. However, those 2 NAs lack the P468H mutation and are antigenically closer to group 1 NAs, confirming the requirement of P468H combined with the S245N and S247T substitutions for this NA antigenic drift (Wan et al., 2019). This observation suggests that NA antigenic drift may sometimes require a combination of amino acid changes rather than single mutations. Wang and colleagues have very elegantly demonstrated that N2 antigenic evolution is constrained by charge balances in an antigenic region, which can partially explain the observation that NA amino acids can undergo co-evolution to result in antigenic drift (Wang et al., 2021). Interestingly, NAs originated from H3N2 variants classified in phylogenetic group 4 were antigenically closer to N2s belonging to phylogenetic group 1. This suggest that the antigenic drift of recent NAs from group 2 increases the antigenic distance from the N2s of swine origin.

Noteworthy, the analysis of the sera obtained from ferrets after infection shows that, despite the lower homologous anti-NA titers, the inhibition pattern evaluated against 5 distinct H6N2s (composed of at least one representative of each antigenic group) is in accordance with the response obtained after boosting with recombinant protein. This suggests that a boost with homologous recombinant soluble tetrameric NA does not alter the breadth of the antibody response induced by H1N2 infection only.

NAI was also determined against 7 distinct H6N2 reassortant viruses using a panel of 43 anti-N2 NA sera raised in mice by immunization with adjuvanted NACHOs. The observed NAI responses again allowed clustering of the N2 NAs into 4 main antigenic groups, which correlates with data obtained in ferrets. Studies in mice can thus offer a convenient alternative to the use of ferrets in future NA antigenicity studies. This data also supports the hypothesis that there may be universal rules of immunogen antigenicity and immunogenicity (Altman et al., 2015).

The determined KM values of the H6N2 viruses used here, indicates that substrate affinity tends to vary according to the phylogenetic relatedness of N2s. The N-glycosylation at position 245 was previously described to be responsible for reducing N2 affinity for its substrate (Powell and Pekosz, 2020). Unlike group 2, which is primarily characterized by the presence of a N-glycan at position 245, NAs from phylogenetic group 4 have a low KM, indicating an increased affinity for substrate. This result suggests that phylogenetic drift can be accompanied by changes in NA enzymatic properties. Curiously, the Ala15, Wis15, and Kan17, despite having antigenic properties closer to group 1, still have KM values that are comparable to those of group 2 N2 NAs.

After the determination of NA antigenic distances, we used random forest modelling to estimate the impact of amino acid changes on antigenic distances. We obtained a strong correlation of predicted versus observed antigenic distances. However, it is important to note that the training dataset was based on the 102 mutations present in our panel, limiting the predictions for other positions. Furthermore, random forest enabled the classification of the variable amino acids according to importance scores. Also, our modelling does not consider that substitution by other amino acids can have a distinct impact on the antigenic distance. As a consequence, predictions based on the model could underestimate the importance of a particular amino acid residue substitution in some cases. It is important to highlight that co-occurring substitutions in our panel (the ones present in the main branches of the phylogenetic tree) cannot be individually assessed by association analysis or the random forest model. The individual weight of those mutation on NA drift thus remains to be experimentally demonstrated. Interestingly, the modelling could put in evidence the gradual drift of recent N2s which were not present in the dataset (A/Hong Kong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021).

It has been reported that an N-glycosylation site at position 329 combined with E344 in NA from human H3N2 viruses from 2007 to 2013 was gradually lost in later H3N2 viruses (Ge et al., 2022). This loss of an N-glycosylation site at position 329 combined with an E344K substitution was associated with a change in NAI reactivity in ferret sera. Three N2 NAs in our panel, derived from A/Victoria/361/2011, A/Hong_Kong/3089/2017, and A/Tennessee/18/2017, lack this N-glycosylation motif. The E344K substitution is present in three other NAs, derived from A/Nagano/2153/2017, A/Minnesota/11/2010, and A/Indiana/08/2011. The importance of those mutations is among the lowest ones predicted by our modeling. However, the differences in NAI reported by Ge et al. are very modest (less than twofold). The experimental variability in our study potentially limits the identification of substitutions with a subtle impact on NAI.

The protective potential of NA antigen against influenza has been underscored by multiple studies. It is likely that NA will be included as a standard component of next generation influenza vaccines. The difficulties of producing stable tetrameric NA may be overcome in the future thanks to protein design or with the advent of mRNA-based vaccines (Catani et al., 2022; Chivukula et al., 2021; Ellis et al., 2022). The antigenic modelling associated with sequence surveillance may increase our predictive ability, highlighting major potential drifts and support NA antigen designs that may elicit broader NAI responses compared to natural NA.

In summary, our study shows that immunization with recombinant N2 NA immunogen can boost anti-NA serum antibody responses in ferrets that were primed by infection with an influenza virus that carries the homologous NA. At least 4 major antigenic groups could be distinguished in N2 NA derived from human H3N2 viruses that were isolated between 2009 to 2017, and those antigenic groups predominantly correspond with the phylogenetic relatedness of the N2 head domains. In addition, the amino acid substitutions in close proximity of the NA catalytic pocket are more likely to affect NAI patterns.

Material and Methods

Rescue of H1N2 and H6N2 reassortants

All reassortant H1N2 and H6N2 viruses included in the study were generated by reverse genetics, with each reassortant expressing the targeted NA antigen, internal genes from A/Puerto Rico/8/1934 H1N1 (PR8), and the HA from PR8 H1N1 or A/mallard/Sweden/81/2002 H6N1, respectively. Briefly, HA and NA segments including non-coding regions were generated by custom gene synthesis (Geneart AG), while PR8 internal segments were derived from a viral isolate as previously published (Marsh and Tannock, 2005). All segments were cloned into a bi-directional transcription plasmid derived from pUC57 (Genscript) including polymerase (Pol) I and Pol II promoters, as previously described (Hoffmann et al., 2000), and the entire set of 8 plasmids was used to transfect 293FT cells (Thermo Fisher Scientific) using Lipofectamine 2000 CD (Thermo Fisher Scientific). Twenty-four hours after transfection, Madin-Darby Canine Kidney Cells (MDCK-) ATL cells (ATCC) were added to the transfected cells in the presence of TPCK-treated trypsin (Sigma) to allow influenza virus propagation. Cell culture supernatants containing influenza virus were harvested 7 days post-MDCK addition and blindly passaged in 8–10-day-old embryonated chicken eggs (Charles River Laboratories, Inc.). Inoculated eggs were incubated at 37°C for 48 h, then cooled to 4°C for 12 h, prior to allantoic fluid harvest and clarification by low-speed centrifugation (3,000 rpm, 20 min). For H6N2 viruses only, high-yield stocks were generated by an additional passage in eggs as described above. Virus titers were determined by plaque assay on MDCK cells.

Design and production of recombinant proteins

The coding information of the NACHO proteins was cloned under the transcriptional control of the CMV promoter in the pCDNA3.4 plasmid with a CD5 secretion signal, an amino-terminal His-tag and the thrombin cleavage signal followed by tetrabrachion-NA. NACHOs were expressed in a mammalian cell culture system as previously described with modifications (Catani et al., 2022).

Animal experiments

The ferret immunization experiment was conducted at Bioqual, Inc., MD (USA). The experiments were approved by the Institutional Animal Care and Use Committee of the Bioqual (reference number 17-021.6). Outbred 18-22 weeks old naïve male and female Fitch ferrets were obtained from Triple F, PA (USA) and housed under specified pathogen-free conditions with food and water ad libitum. Ferrets were primed on day 0 by intranasal challenge with influenza H1N2 reassortant virus (1,000µl/dose, split evenly between nostrils). Four weeks after initial infection (day 28), ferrets were immunized with 45 μg of NACHO (500µl/dose, intramuscularly, split evenly between hind thighs) corresponding to homologous NA used in day 0 prime. Alternatively, and due to low yield of selected recombinant NACHOs, ferrets were boosted on day 28 with 15 μg of homologous NACHO (A/Perth/16/2009nib-64, A/Tasmania/1018/2015, A/Utah/11/2011, and A/West Virginia/17/2012) or were inoculated with the same H1N2 reassortant virus used in the prime (A/Estonia/91621/2015 and A/Stockholm/15/2014). Sera from bleeds two weeks after the prime (day 14) and booster vaccination (day 42) were tested for immunogenicity and seroconversion in serological assays.

The mouse immunization experiments were conducted according to the Belgian legislation (Belgian Law 14/08/1986 and Belgium Royal Decree 06/04/2010) and European legislation on protection of animals used for scientific purposes (EU directives 2010/63/EU and 86/609/EEC). Experimental protocols were all approved by the Ethics Committee of the Vlaams Instituut voor Biotechnologie (VIB), Ghent University, Faculty of Science (EC2019-035). Female BALB/c mice, aged 6–7 weeks, were purchased from Charles River. The mice were housed in a specified pathogen-free animal house with food and water ad libitum. Animals were immunized intramuscularly in the right quadriceps. Immunization was performed with 50 µl containing 1 µg of recombinant protein (tetrabrachion stabilized NAs, referred to as NACHOs). All, except A/West Virginia/17/2012 NACHOs were used as immunogens in mice. All immunizations were adjuvanted with a 1:1 volume of AF03 (25 μl of antigen in PBS + 25 μl of AF03 per dose) (Caillet et al., 2010).

Two or three weeks after the boost immunization, mice were humanely sacrificed with an anaesthetic overdose of Nembutal (6 mg per mouse) and bled retro-orbitally. The obtained blood samples were incubated for 30 min at 37°C to allow clotting, which was followed by centrifugation at 10,500 g for 5 minutes. The supernatant (the serum) was recovered and submitted to a second centrifugation at 10,500 g for 5 minutes. Cleared sera were stored at −20°C before use in serological assays.

ELISA

Anti-tetNA IgG levels in mouse and ferret sera were determined by capture ELISA using tetNAs in PierceTM nickel-coated plates (cat. # 15442). Recombinant proteins were diluted in DPBS (Life technologies cat. # 14040-182) (tetNA was diluted to 0.5 µg/ml). Then, 50 µl of the coating antigen solutions was added to each well and the plate was incubated at room temperature for 1 h on a shaking platform. The wells of the plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and blocked for 1h with 1% BSA (for the mouse sera) or 10% milk (for the ferret sera) in DPBS. After blocking, the wells of the plates were washed once with PBS-T and incubated with a 3-fold serial dilution, starting at a 1/100 dilution, of serum in DPBS, 0.5% BSA, 0.05% tween-20 for 2h at room temperature on a shaking platform. The wells of the plates were then washed 5 times with PBS-T and incubated with a 1:5000 dilution of anti-mouse IgG-HRP (GE healthcare cat. # NA931-1ml), or with a 1:100000 dilution of anti-ferret IgG-HRP (Abcam cat. # 112770) in DPBS, 0.5% BSA, 0.05% Tween-20. The 3,3’,5,5’-tetramethylbenzidine (TMB) substrate (BD cat. # 555214) was added after 3 washes with PBS-T and the reaction was stopped after 5 min by addition of 50 µl of 1 M H2SO4. The optical density (O.D.) in each well was determined at 450 nm and, as a reference, 655 nm using an iMark™ Microplate Absorbance Reader (Bio-rad). The end point titer was determined for each serum sample by scoring the dilution that resulted in an O.D. that was equal to or 2 times higher than the background O.D. obtained from pre-immune control sera dilution series.

Enzyme-linked lectin assay to determine neuraminidase inhibition titers

Fetuin (Sigma cat. # F3385) was diluted into coating buffer (KPL cat. # 50-84-01) to a concentration of 25 µg/ml and 50 µl was added to the wells of Nunc MaxiSorp™ plates (Thermofisher cat. # 44-2404-21), which were incubated overnight at 4°C. The coated plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and incubated overnight with 60 μl of a dilution of H6N2 and 60 µl of 2-fold serial dilution of serum, starting at a 1/20 dilution, in sample buffer (1X MES VWR cat. # AAJ61979-AP: 20mM CaCl2, 1% BSA, 0.5% Tween-20). These dilutions of the H6Nx viruses correspond to the 70% maximum activity of NA from the respective viruses as determined in the ELLA assay. Fetuin-coated plates were then washed 3 times with PBS-T and incubated for 1 h with a solution of PNA-HRP (cat. # L6135-1MG, Sigma) at 5 μg/ml in conjugate diluent (MES pH 6.5, 20 mM CaCl2, 1% BSA). The plates were washed 3 times with PBS-T, TMB substrate was added and then the plates were incubated for 5 minutes before the reaction was stopped by the addition of 50 µl of 1 M H2SO4. The optical density was measured at 450 nm and, as a reference, 655 nm in an iMark™ Microplate Absorbance Reader (Bio-rad). Half maximum inhibitory concentrations (IC50) values were determined by non-linear regression analysis (GraphPad Prism software).

NA Enzyme Kinetics Assay

The enzymatic activity of NACHOs and H6N2s was determined with the fluorogenic small substrate 4-methylumbelliferyl)-α-d-N-acetylneuraminic acid (MUNANA) as previously described (Marathe et al., 2013). First H6N2 dilutions were tested to ensure that MUNANA conversion into 4-methylumbelliferone (4-MU) generated fluorescence over time at a linear ratio. Ten microliters of 1 mM of MUNANA (Sigma cat # M8639) in 200 mM sodium acetate buffer (pH 6.5) containing 2 mM CaCl2 and 1% butanol was incubated with 40 µl of PBS solution containing viral dilution. Conversion to 4-MU was monitored every 2 min for 1 h using a BMG Fluostar OPTIMA reader (excitation at 365 nm and emission determined at 450 nm). Second, using the selected dilution of virus, the assay was repeated with dilutions of MUNANA ranging from 2.5 to 100 µM. A standard curve with 800 to 25 pmoles/well of 4-MU (Sigma cat # M1508) was used to extrapolate the molar conversion of MUNANA. KM was calculated using GraphPad Prism version 8.4.3.

Antigenic cartography

Antigenic cartography was performed as described by Smith et al. (Smith et al., 2004) using the R CRAN package (https://CRAN.R-project.org/package=Racmacs).

Association study

For the association analysis, under the assumption of homogeneous genetic relatedness among the selected H3N2 virus strains, a simple linear regression model for mutation-trait association analysis was fitted to the data for each H6N2 separately. The significance values were calculated by comparing IC50 values of sera that were classified into 2 groups according to the presence or absence of variation in the amino acid of reference Tex12 sequence.

Correlations with distances from the NA catalytic pocket

Tau values were determined by correlating distances with significance values obtained from the association analysis or the importance scores obtained from machine learning. A similar method was described by Meyer et al. to define geometrical constraints of HA evolution (Meyer and Wilke, 2015). Briefly, the relative distances of amino acids were calculated based on the structure of A/Perth/2009 (H3N2) NA (PDB: 6BR6) (Hadházi et al., 2018). The distances were calculated from each Cα to every other Cα. The distances from every residue to the 98 variable residues (4 of the 102 variable amino acids are localized near the end of the stalk and are not resolved in the crystal structure) residues of the that are present in the panel of N2s and the 6BR6 structure were then correlated with the significance values that were obtained in the association analysis, or importance scores obtained in the machine learning. Correlation coefficients were visualized on the structure of N2 NA derived from pandemic A/RI/5+/1957 H2N2 virus (PDB: 4H53)(Vavricka et al., 2013) using Chimera 1.14.

Random Forest algorithm

The Random Forest algorithm was adapted from (Yao et al., 2017). Briefly, the non-conserved positions in the multiple sequence alignment of N2 head domain amino acid sequences of all tested N2s were selected. The feature matrix (X) was created by assessing matches/mismatches on each non-conserved position for each pair of N2 head domain amino acid sequences as: where Xi represents the non-conserved positions in the multiple sequence alignment

m is the number of non-conserved positions (102 in total)

Vi represents the N2 head domain sequence for each tested NA

n is the number of tested sera (43 in total)

0 represents a match, 1 represents a mismatch in the alignment

The response vector (Y) is composed of the antigenic distances for each pair of tested sera. These antigenic distances were determined based on NAI, similarly to the method used to determine HI based immunogenic distances that was described by Cai et al. In summary, the distances between sera was calculated as , which represent average distances between 2 serum samples among all tested H6N2s (Cai et al., 2012). We applied the random forest function from the randomForest package in R to construct the model. We set the bootstrapping (tree) number to 1500 and the number of features to 25 for each tree. Accuracy was determined as the root-mean-square error (RMSE), calculated using the RMSE function from the yardstick package in R.

Additionally, predicted antigenic distances of the out-of-bag samples were compared to the observed distances by fitting a linear model and calculating the R2 value. To calculate the importance of each feature (Xi), the following metrics were calculated using the measure importance function of the randomForestExplainer package in R:

– mean minimal depth: mean of the depths of the node in each tree that splits on this feature and is the closest to the root of the tree (minimal depth). The lower, the more important the feature since a low depth means that many observations are divided into groups based on this feature.

– mse decrease: mean decrease of accuracy in predictions on out of bag samples when the feature is excluded from the model. The higher, the more important the feature.

– number of nodes: total number of nodes in the forest that use the feature for splitting. The higher, the more important the feature.

The data and code used for the generation of the rf model is available at https://github.com/SaelensLAB/RF.

eLife Assessment

This study presents valuable data on the antigenic properties of neuraminidase proteins of human A/H3N2 influenza viruses sampled between 2009 and 2017. The antigenic properties are found to be generally concordant with genetic groups. Additional analysis have strengthened the revised manuscript, and the evidence supporting the claims is solid.

Reviewer #1 (Public Review):

Summary

The authors investigated the antigenic diversity of recent (2009-2017) A/H3N2 influenza neuraminidases (NAs), the second major antigenic protein after haemagglutinin. They used 27 viruses and 43 ferret sera and performed NA inhibition. This work was supported by a subset of mouse sera. Clustering analysis determined 4 antigenic clusters, mostly in concordance with the genetic groupings. Association analysis was used to estimate important amino acid positions, which were shown to be more likely close to the catalytic site. Antigenic distances were calculated and a random forest model used to determine potential important sites.

This revision has addressed many of my concerns of inconsistencies in the methods, results and presentation. There are still some remaining weaknesses in the computational work.

Strengths

(1) The data cover recent NA evolution and a substantial number (43) of ferret (and mouse) sera were generated and titrated against 27 viruses. This is laborious experimental work and is the largest publicly available neuraminidase inhibition dataset that I am aware of. As such, it will prove a useful resource for the influenza community.

(2) A variety of computational methods were used to analyse the data, which give a rounded picture of the antigenic and genetic relationships and link between sequence, structure and phenotype.

(3) Issues raised in the previous review have been thoroughly addressed.

Weaknesses

(1) Some inconsistencies and missing data in experimental methods

Two ferret sera were boosted with H1N2, while recombinant NA protein for the others. This, and the underlying reason, are clearly explained in the manuscript. The authors note that boosting with live virus did not increase titres. Additionally, one homologous serum (A/Kansas/14/2017) was not generated, although this would not necessarily have impacted the results.

(2) Inconsistency in experimental results

Clustering of the NA inhibition results identifies three viruses which do not cluster with their phylogenetic group. Again this is clearly pointed out in the paper and is consistent with the two replicate ferret sera. Additionally, A/Kansas/14/2017 is in a different cluster based on the antigenic cartography vs the clustering of the titres

(3) Antigenic cartography plot would benefit from documentation of the parameters and supporting analyses

a. The number of optimisations used

b. The final stress and the difference between the stress of the lowest few (e.g. 5) optimisations, or alternatively a graph of the stress of all the optimisations. Information on the stress per titre and per point, and whether any of these were outliers

c. A measure of uncertainty in position (e.g. from bootstrapping)

(4) Random forest

The full dataset was used for the random forest model, including tuning the hyperparameters. It is more robust to have a training and test set to be able to evaluate overfitting (there are 25 features to classify 43 sera).

Reviewer #2 (Public Review):

Summary:

The authors characterized the antigenicity of N2 protein of 43 selected A(H3N2) influenza A viruses isolated from 2009-2017 using ferret and mice immune sera. Four antigenic groups were identified, which the authors claimed to be correlated with their respective phylogenic/ genetic groups. Among 102 amino acids differed by the 44 selected N2 proteins, the authors identified residues that differentiate the antigenicity of the four groups and constructed a machine-learning model that provides antigenic distance estimation. Three recent A(H3N2) vaccine strains were tested in the model but there was no experimental data to confirm the model prediction results.

Strengths:

This study used N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 and generated corresponding panels of ferret and mouse sera to react with the selected strains. The amount of experimental data for N2 antigenicity characterization is large enough for model building.

Weaknesses:

The main weakness is that the strategy of selecting 43 A(H3N2) viruses from 2009-2017 was not explained. It is not clear if they represent the overall genetic diversity of human A(H3N2) viruses circulating during this time. In response to the reviewer's comment, the authors have provided a N2 phylogenetic tree using180 randomly selected N2 sequences from human A(H3N2) viruses from 2009-2017. While the 43 strains seems to scatter across the N2 tree, the four antigenic groups described by the author did not correlated with their respective phylogenic/ genetic groups as shown in Fig. 2. The authors should show the N2 phylogenic tree together with Fig. 2 and discuss the discrepancy observed.

The second weakness is the use of double-immune ferret sera (post-infection plus immunization with recombinant NA protein) or mouse sera (immunized twice with recombinant NA protein) to characterize the antigenicity of the selected A(H3N2) viruses. Conventionally, NA antigenicity is characterized using ferret sera after a single infection. Repeated influenza exposure in ferrets has been shown to enhance antibody binding affinity and may affect the cross-reactivity to heterologous strains (PMID: 29672713). The increased cross-reactivity is supported by the NAI titers shown in Table S3, as many of the double immune ferret sera showed the highest reactivity not against its own homologous virus but to heterologous strains. In response to the reviewer's comment, the authors agreed the use of double-immune ferret sera may be a limitation of the study. It would be helpful if the authors can discuss the potential effect on the use of double-immune ferret sera in antigenicity characterization in the manuscript.

Another weakness is that the authors used the newly constructed a model to predict antigenic distance of three recent A(H3N2) viruses but there is no experimental data to validate their prediction (eg. if these viruses are indeed antigenically deviating from group 2 strains as concluded by the authors). In response to the comment, the authors have taken two strains out of the dataset and use them for validation. The results is shown as Fig. R7. However, it may be useful to include this in the main manuscript to support the validity of the model.

Reviewer #3 (Public Review):

Summary:

This paper by Portela Catani et al examines the antigenic relationships (measured using monotypic ferret and mouse sera) across a panel of N2 genes from the past 14 years, along with the underlying sequence differences and phylogenetic relationships. This is a highly significant topic given the recent increased appreciation of the importance of NA as a vaccine target, and the relative lack of information about NA antigenic evolution compared with what is known about HA. Thus, these data will be of interest to those studying the antigenic evolution of influenza viruses. The methods used are generally quite sound, though there are a few addressable concerns that limit the confidence with which conclusions can be drawn from the data/analyses.

Strengths:

-The significance of the work, and the (general) soundness of the methods.

-Explicit comparison of results obtained with mouse and ferret sera

Weaknesses:

- Approach for assessing influence of individual polymorphisms on antigenicity does not account for potential effects of epistasis (this point is acknowledged by the authors).

- Machine learning analyses neither experimentally validated nor shown to be better than simple, phylogenetic-based inference.

Author response:

The following is the authors’ response to the previous reviews.

eLife assessment

This study presents valuable data on the antigenic properties of neuraminidase proteins of human A/H3N2 influenza viruses sampled between 2009 and 2017. The antigenic properties are found to be generally concordant with genetic groups. Additional analysis have strengthened the revised manuscript, and the evidence supporting the claims is solid.

Public Reviews:

Reviewer #1 (Public Review):

Summary

The authors investigated the antigenic diversity of recent (2009-2017) A/H3N2 influenza neuraminidases (NAs), the second major antigenic protein after haemagglutinin. They used 27 viruses and 43 ferret sera and performed NA inhibition. This work was supported by a subset of mouse sera. Clustering analysis determined 4 antigenic clusters, mostly in concordance with the genetic groupings. Association analysis was used to estimate important amino acid positions, which were shown to be more likely close to the catalytic site. Antigenic distances were calculated and a random forest model used to determine potential important sites.

This revision has addressed many of my concerns of inconsistencies in the methods, results and presentation. There are still some remaining weaknesses in the computational work.

Strengths

(1) The data cover recent NA evolution and a substantial number (43) of ferret (and mouse) sera were generated and titrated against 27 viruses. This is laborious experimental work and is the largest publicly available neuraminidase inhibition dataset that I am aware of. As such, it will prove a useful resource for the influenza community.

(2) A variety of computational methods were used to analyse the data, which give a rounded picture of the antigenic and genetic relationships and link between sequence, structure and phenotype.

(3) Issues raised in the previous review have been thoroughly addressed.

Weaknesses

(1). Some inconsistencies and missing data in experimental methods Two ferret sera were boosted with H1N2, while recombinant NA protein for the others. This, and the underlying reason, are clearly explained in the manuscript. The authors note that boosting with live virus did not increase titres. Additionally, one homologous serum (A/Kansas/14/2017) was not generated, although this would not necessarily have impacted the results.

We agree with the reviewer and this point was addressed in the previous rebuttal.

(2) Inconsistency in experimental results

Clustering of the NA inhibition results identifies three viruses which do not cluster with their phylogenetic group. Again this is clearly pointed out in the paper and is consistent with the two replicate ferret sera. Additionally, A/Kansas/14/2017 is in a different cluster based on the antigenic cartography vs the clustering of the titres

We agree with the reviewer and this point was addressed in the previous rebuttal.

(3) Antigenic cartography plot would benefit from documentation of the parameters and supporting analyses

a. The number of optimisations used

We used 500 optimizations. This information is now included in the Methods section.

b. The final stress and the difference between the stress of the lowest few (e.g. 5) optimisations, or alternatively a graph of the stress of all the optimisations. Information on the stress per titre and per point, and whether any of these were outliers

The stress was obtained from 1, 5, 500, or even 5000 optimizations (resulting in stress values of respectively, 1366.47, 1366.47, 2908.60, and 3031.41). Besides limited variation or non-conversion of the stress values after optimization, the obtained maps were consistent in multiple runs. The map was obtained keeping the best optimization (stress value 1366.47, selected using the keepBestOptimization() function).

The stress per point is presented in the heat map below.

The heat map indicates stress per serum (x-axis) and strain (y-axis) in blue to red scale.

c. A measure of uncertainty in position (e.g. from bootstrapping)

Bootstrap was performed using 1000 repeats and 100 optimizations per repeat. The uncertainty is represented in the blob plot below.

(4) Random forest

The full dataset was used for the random forest model, including tuning the hyperparameters. It is more robust to have a training and test set to be able to evaluate overfitting (there are 25 features to classify 43 sera).

Explicit cross validation is not necessary for random forests as the out of bag process with multiple trees implicitly covers cross validation. In the random forest function in R this is done by setting the mtry argument (number of variables randomly sampled as candidates at each split). R samples variables with replacement (the same variable can be sampled multiple times) of the candidates from the training set. RF will then automatically take the data that is not selected as candidates as test set. Overfit may happen when all data is used for training but the RF method implicitly does use a test set and does not use all data for training.

Code:

rf <- randomForest(X,y=Y,ntree=1500,mtry=25,keep.forest=TRUE,importance=TRUE)

Reviewer #2 (Public Review):

Summary:

The authors characterized the antigenicity of N2 protein of 43 selected A(H3N2) influenza A viruses isolated from 2009-2017 using ferret and mice immune sera. Four antigenic groups were identified, which the authors claimed to be correlated with their respective phylogenic/ genetic groups. Among 102 amino acids differed by the 44 selected N2 proteins, the authors identified residues that differentiate the antigenicity of the four groups and constructed a machine-learning model that provides antigenic distance estimation. Three recent A(H3N2) vaccine strains were tested in the model but there was no experimental data to confirm the model prediction results.

Strengths:

This study used N2 protein of 44 selected A(H3N2) influenza A viruses isolated from 2009-2017 and generated corresponding panels of ferret and mouse sera to react with the selected strains. The amount of experimental data for N2 antigenicity characterization is large enough for model building.

Weaknesses:

The main weakness is that the strategy of selecting 43 A(H3N2) viruses from 2009-2017 was not explained. It is not clear if they represent the overall genetic diversity of human A(H3N2) viruses circulating during this time. In response to the reviewer's comment, the authors have provided a N2 phylogenetic tree using180 randomly selected N2 sequences from human A(H3N2) viruses from 2009-2017. While the 43 strains seems to scatter across the N2 tree, the four antigenic groups described by the author did not correlated with their respective phylogenic/ genetic groups as shown in Fig. 2. The authors should show the N2 phylogenic tree together with Fig. 2 and discuss the discrepancy observed.

The discrepancies between the provided N2 phylogenetic tree using 180 selected N2 sequences was primarily due to visualization. In the tree presented in Figure 2 the phylogeny was ordered according to branch length in a decreasing way. Further, the tree represented in the rebuttal was built with PhyML 3.0 using JTT substitution model, while the tree in figure 2 was build in CLC Workbench 21.0.5 using Bishop-Friday substitution model. The tree below was built using the same methodology as Figure 2, including branch size ordering. No discrepancies are observed.

Phylogenetic tree representing relatedness of N2 head domain. N2 NA sequences were ordered according to the branch length and phylogenetic clusters are colored as follows: G1: orange, G2: green, G3: blue, and G4: purple. NA sequences that were retained in the breadth panel are named according to the corresponding H3N2 influenza viruses. The other NA sequences are coded.

The second weakness is the use of double-immune ferret sera (post-infection plus immunization with recombinant NA protein) or mouse sera (immunized twice with recombinant NA protein) to characterize the antigenicity of the selected A(H3N2) viruses. Conventionally, NA antigenicity is characterized using ferret sera after a single infection. Repeated influenza exposure in ferrets has been shown to enhance antibody binding affinity and may affect the cross-reactivity to heterologous strains (PMID: 29672713). The increased cross-reactivity is supported by the NAI titers shown in Table S3, as many of the double immune ferret sera showed the highest reactivity not against its own homologous virus but to heterologous strains. In response to the reviewer's comment, the authors agreed the use of double-immune ferret sera may be a limitation of the study. It would be helpful if the authors can discuss the potential effect on the use of double-immune ferret sera in antigenicity characterization in the manuscript.

Our study was designed to understand the breadth of the anti-NA response after the incorporation of NA as a vaccine antigens. Our data does not allow to conclude whether increased breadth of protection is merely due to increased antibody titers or whether an NA boost immunization was able to induce antibody responses against epitopes that were not previously recognized by primary response to infection. However, we now mention this possibility in the discussion and cite Kosikova et al. CID 2018, in this context.

Another weakness is that the authors used the newly constructed a model to predict antigenic distance of three recent A(H3N2) viruses but there is no experimental data to validate their prediction (eg. if these viruses are indeed antigenically deviating from group 2 strains as concluded by the authors). In response to the comment, the authors have taken two strains out of the dataset and use them for validation. The results is shown as Fig. R7. However, it may be useful to include this in the main manuscript to support the validity of the model.

The removal of 2 strains was performed to illustrate the predictive performance of the RF modeling. However, Random Forest does not require cross-validation. The reason is that RF modeling already uses an out-of-bag evaluation which, in short, consists of using only a fraction of the data for the creation of the decision trees (2/3 of the data), obviating the need for a set aside the test set:

“…In each bootstrap training set, about one-third of the instances are left out. Therefore, the out-of-bag estimates are based on combining only about one- third as many classifiers as in the ongoing main combination. Since the error rate decreases as the number of combinations increases, the out-of-bag estimates will tend to overestimate the current error rate. To get unbiased out-of-bag estimates, it is necessary to run past the point where the test set error converges. But unlike cross-validation, where bias is present but its extent unknown, the out-of-bag estimates are unbiased…” from https://www.stat.berkeley.edu/%7Ebreiman/randomforest2001.pdf

Reviewer #3 (Public Review):

Summary:

This paper by Portela Catani et al examines the antigenic relationships (measured using monotypic ferret and mouse sera) across a panel of N2 genes from the past 14 years, along with the underlying sequence differences and phylogenetic relationships. This is a highly significant topic given the recent increased appreciation of the importance of NA as a vaccine target, and the relative lack of information about NA antigenic evolution compared with what is known about HA. Thus, these data will be of interest to those studying the antigenic evolution of influenza viruses. The methods used are generally quite sound, though there are a few addressable concerns that limit the confidence with which conclusions can be drawn from the data/analyses.

Strengths:

The significance of the work, and the (general) soundness of the methods. -Explicit comparison of results obtained with mouse and ferret sera

Weaknesses:

Approach for assessing influence of individual polymorphisms on antigenicity does not account for potential effects of epistasis (this point is acknowledged by the authors).

We agree with the reviewer and this point was addressed in the previous rebuttal.

Machine learning analyses neither experimentally validated nor shown to be better than simple, phylogenetic-based inference.

We respectfully disagree with the reviewer. This point was addressed in the previous rebuttal as follows.

This is a valid remark and indeed we have found a clear correlation between NAI cross reactivity and phylogenetic relatedness. However, besides achieving good prediction of the experimental data (as shown in Figure 5 and in FigureR7), machine Learning analysis has the potential to rank or indicate major antigenic divergences based on available sequences before it has consolidated as new clade. ML can also support the selection and design of broader reactive antigens. “

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

(1) Discuss the discrepancy between Fig. 2 and the newly constructed N2 phylogenetic tree with 180 randomly selected N2 sequences of A(H3N2) viruses from 2009-2017. Specifically please explain the antigenic vs. phylogenetic relationship observed in Fig. 2 was not observed in the large N2 phylogenetic tree.

Discrepancies were due to different method and visualization. A new tree was provided.

(2) Include a sentence to discuss the potential effect on the use of double-immune ferret sera in antigenic characterization.

We prefer not to speculate on this.

(3) Include the results of the exercise run (with the use of Swe17 and HK17) in the manuscript as a way to validate the model.

The exercise was performed to illustrate predictive potential of the RF modeling to the reviewer. However, cross-validation is not a usual requirement for random forest, since it uses out-of-bag calculations. We prefer to not include the exercise runs within the main manuscript.

Version 3

Abstract

Abstract

Human H3N2 influenza viruses are subject to rapid antigenic evolution which translates into frequent updates of the composition of seasonal influenza vaccines. Despite these updates, the effectiveness of influenza vaccines against H3N2-associated disease is suboptimal. Seasonal influenza vaccines primarily induce hemagglutinin-specific antibody responses. However, antibodies directed against influenza neuraminidase (NA) also contribute to protection. Here, we analyzed the antigenic diversity of a panel of N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. The antigenic breadth of these NAs was determined based on the NA inhibition (NAI) of a broad panel of ferret and mouse immune sera that were raised by infection and recombinant N2 NA immunization. This assessment allowed us to distinguish at least 4 antigenic groups in the N2 NAs derived from human H3N2 viruses that circulated between 2009 and 2017. Computational analysis further revealed that the amino acid residues in N2 NA that have a major impact on susceptibility to NAI by immune sera are in proximity of the catalytic site. Finally, a machine learning method was developed that allowed to accurately predict the impact of mutations that are present in our N2 NA panel on NAI. These findings have important implications for the renewed interest to develop improved influenza vaccines based on the inclusion of a protective NA antigen formulation.

Introduction

Neuraminidase (NA) and hemagglutinin (HA) constitute the two major antigenic proteins at the surface of influenza A and B virions (Uyeki et al., 2022). While HA mediates binding to sialic acids and membrane fusion, NA catalyses the removal of sialic acid residues from sialylated glycoproteins and glycolipids. This complementary role of HA and NA enables influenza virions to penetrate the mucus barrier that lines the respiratory epithelium and newly formed virions to be released from the infected cell (de Vries et al., 2020; Matrosovich et al., 2004; Palese et al., 1974). Immunity against influenza is associated with the presence of anti-HA antibodies in serum, as determined by a hemagglutination inhibition assay. In line with this, the composition of licensed influenza vaccines is primarily focused on the antigenic characteristics of HA as the protective vaccine antigen (Cox, 2013). Multiple studies, however, have underscored the protective potential of anti-NA immune responses as determined by NA inhibition (NAI) assays (Couch et al., 2013; Maier et al., 2019; Memoli et al., 2016; Monto and Kendal, 1973; Monto et al., 2015; Ng et al., 2019; Weiss et al., 2019). There is also evidence that pre-existing anti-N2 immunity induced by prior exposure to H2N2 viruses that circulated in the population from 1957 to 1968, tempered the impact of the H3N2 pandemic virus that emerged in 1968 (Eickhoff and Meiklejohn, 1969; Jester et al., 2020; Monto and Kendal, 1973; Viboud et al., 2005). Further, the pharmacological inhibition of NA activity is a well-established therapeutic strategy to try to reduce influenza severity and duration (Leneva et al., 2000; Yen et al., 2005). In addition, antibodies with NAI activity can protect against influenza virus infection in pre-clinical models (Chen et al., 2018; Deroo et al., 1996; Rosu et al., 2022; Wan et al., 2018; Wohlbold et al., 2017).

Since its appearance in 1968, human H3N2 viruses have been evolving faster than the co-circulating H1N1 and influenza B viruses (Bedford et al., 2014; Potter et al., 2019). The overall Influenza virus vaccine effectiveness varies between seasons and, despite frequent updates of the H3N2 HA component, the effectiveness of influenza vaccines against H3N2-associated disease ranges between 5 and 39%, which is lower than the effectiveness against H1N1 and influenza B (Belongia et al., 2016). In this context the use of NA as a complementary vaccine antigen might improve influenza vaccine effectiveness, for example by compensating for HA-based antigenic mismatches in circulating influenza virus strains (Krammer et al., 2018).

Phylogenetic analyses of HA and NA genes of human H3N2 viruses revealed the asynchronous evolution of the 2 proteins. In addition, the evolutionary rate as determined by the accumulation of mutations over time, is estimated to be 1.5 times lower for NA than for HA (Westgeest et al., 2012). Despite the strong evidence that immune responses directed against NA can contribute to protection against influenza, a comprehensive insight in the antigenic breadth and evolution of the H3N2 NA antigen is lacking (Nachbagauer et al., 2017; Rajendran et al., 2017; Smith et al., 2004; Westgeest et al., 2012).

Here, we determined the antigenic landscape of N2 NAs derived from human H3N2 strains that were isolated between 2009 and 2017. This analysis allowed the identification of at least 4 N2 NA antigenic groups and to estimate the weight of amino acid changes on their contribution to NA antigenic drift.

Results

Generation of H3N2-derived NA immune sera in ferrets and mice

Our aim was to determine the antigenic relationship between NAs derived from human H3N2 viruses that circulated during nearly one decade. We first selected 43 H3N2 neuraminidase protein sequences obtained via GISAID, derived from viruses that were isolated between 2009 and 2017 (Elbe and Buckland-Merrett, 2017). Hierarchical clustering with minimax linkage was employed to ensure the set was representative of overall protein sequence diversity (Bien and Tibshirani, 2011). Next, 43 recombinant H1N2 viruses were generated that carried the internal and HA genes from PR8 virus and the NA genes of the selected H3N2 viruses. In addition, the corresponding tetrabrachion-stabilized recombinant N2 NA proteins were produced in Chinese hamster ovary (CHO) cells and purified, and are referred to as NACHOs (Catani et al., 2022). All purified NACHOs were tetrameric based on SEC-MALS analysis. The enzymatic activity of the NACHOs was determined based on the hydrolysis of the fluorogenic small molecule substrate 4-methylumbelliferyl-N-acetylneuraminic acid (MUNANA). The NACHOs were active with a specific activity ranging from 0.27 to 27.48 nmole/min/µg, indicating correct folding and assembly of the NA head domains (Table S1).

Immune sera against all 43 H3N2-derived NAs were generated in pairs of ferrets by a primary infection with the respective H1N2 reassortant viruses. This prime by infection was used to avoid the use of an adjuvant upon subsequent boost with the corresponding NACHOs. On day 28 after primary infection, all animals were boosted by intramuscular injection of homologous NACHO, except for the ferrets that had been primed by infection with A/Stockholm/15/2014 and A/Estonia/91621/2015, which were re-inoculated with the same respective H1N2 reassortant viruses. These two pairs of ferrets were not boosted with homologous NACHO because the corresponding NACHOs were not available at the time of the experiment. Blood samples were taken from the ferrets on days 14 and 42 after (primary) infection (Figure 1a). Priming by H1N2 infection resulted in seroconversion against the respective immobilized homologous NACHOs (Figure 1b). Subsequent immunization with recombinant NACHOs boosted the NACHO-specific IgG titers in day 42 serum samples. In contrast, serum samples from the ferrets that had been re-inoculated with H1N2 reassortant carrying NA derived from A/Stockholm/15/2014 or A/Estonia/91621/2015, showed no further increase in titers, presumably because prior exposure with the same H1N2 virus had induced a sterilizing immune response (Figure 1b).

Generation of anti-N2 sera panel from ferrets and mice.

(a) 43 Ferret immunization and blood sampling scheme. The second immunization on day 28 was performed with homologous NACHO except for the ferrets that had been primed with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014, which were infected again on day 28 post prime with the same H1N2 viruses. (b) Homologous NACHO ELISA titer of ferret sera sampled on day 14 after inoculation with H1N2 virus and 2 weeks after the boost with recombinant NA or by re-inoculation with H1N2A/Estonia/91621/2015 or H1N2A/Stockholm/15/2014 (indicated by arrows on the X-axis). (c) 43 BALB/c mouse immunization and blood sampling scheme. (d) Homologous NACHO serum IgG titers in pooled sera from mice that were immunized twice with recombinant NA. Mouse and ferrets representations were created with BioRender.com (2023).

We also generated NA immune sera in BALB/c mice. Groups of 5 mice were immunized two times, with a two-week interval, by intramuscular injection with 1 µg of the respective 43 NACHOs adjuvanted with AF03. Blood was collected from the immunized mice and serum prepared on day 28 after the prime immunization (Figure 1c). NACHO-specific IgG titers in the mouse immune sera, pooled from each group, ranged from 1:8100 to 1:72900 (Figure 1d).

The antigenic profile of N2 neuraminidase

Next, NAI titers were determined in the ferret sera using a panel of H6N2 reassortant viruses in an enzyme-linked lectin assay (ELLA), avoiding interference by anti-H1 HA antibodies. We generated 27 H6N2 influenza viruses with H6 HA derived from A/mallard/Sweden/2002 and NA derived from a subset of the 43 H3N2 viruses, thereby retaining NA sequence diversity and year of H3N2 virus isolation (Table S2).

The 43 NAs used as antigens and 27 NAs as source of NA, could be classified into 4 groups when ordered according to the phylogenetic relatedness of their NA sequence (Figure 2). The conversion rate of MUNANA into 4-methylumbelliferyl and acetylneuraminic acid by these H6N2 viruses revealed a KM that ranged from 20.7 to 144.8 µM as determined by the Michaelis-Menten equation (Figure 2 – figure supplement 1a and Table S2). Interestingly, the KM differs between the distinct phylogenetic groups (Figure 2 – figure supplement 1a). There was a negative correlation between the KM values and the dilution of the H6N2 viruses required to set the 70% maximal NA activity in ELLA (Figure 2 – figure supplement 1b).

Breadth of N2 NA inhibition.

43 Phylogenetic tree based on the N2 NA head domain amino acid sequences and heat-map representing the average of normalized neuraminidase inhibition per H6N2 [log2 (max (NAI/ ELISA homologous titer)/(NAI/ homologous ELISA titer)] determined in ferret sera after the boost (listed vertically). The red-to-blue scale indicates high-to-low NAI observed in ELLA against the H6N2 reassortants (listed at the bottom). UPGMA clustering of H6N2s inhibition profiles are shown on top of the heat map and coloured according to the phylogenetic groups.

Using this panel of 27 H6N2s, we determined the NAI titers in the sera obtained from the 43 pairs of ferrets 2 weeks after the boost. ELLAs were performed in duplicate for all serum samples against one H6N2 target at the same time. The reciprocal NAI titers of the ferret sera paired with H6N2 viruses are shown in Table S3. In general, the NAI titers of the ferret sera were the highest against homologous and phylogenetically closely related H6N2 viruses. In Figure 2, all the tested ferret sera are represented in the rows and the 27 H6N2 viruses against which the sera were evaluated in ELLA are presented in the columns. The serum samples were ordered according to the phylogenetic relatedness of the N2 NA head domain amino acid sequences, showing that these cluster in 4 groups (G1-G4). The ELLA-based IC50 NAI titers were normalized by the homologous ELISA titers and then plotted relative to the maximum NAI titer obtained against the respective H6N2 target viruses. The heatmap thus represents, for each H6N2 virus, the normalized ELLA IC50 titer expressed as log2 of the fold-difference relative to the serum with the highest IC50 value against that H6N2 virus (dark red squares indicate high and blue squares low IC50 values) (Figure 2).

Next, these normalized NAI values were clustered against each of the H6N2s using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA), which is a distance-based method. This clustering allowed to discern at least 4 major antigenic clusters (AC1 to −4) that are depicted at the top of the heat map (Figure 2). It is interesting to note that the H6N2 viruses with NA antigens derived from A/Wisconsin/16/15 and A/Alaska/251/15 do not segregate into AC2, as would be predicted based on the phylogenetic relatedness, but cluster in AC1 based on the UPGMA. However, the overall concordance of the antigenicity with the phylogenetic relatedness, based on the amino acid sequences of the NA head domains, was confirmed by hierarchical clustering with bootstrapping and k-means (Figure 2 – figure supplements 2a and 2b).

In addition, antigenic cartography analysis, based on the method described by Smith et al. (Smith et al., 2004), resulted in a antigenic map of the N2 NAs that was in line with the 4 proposed antigenic clusters (Figure 2 – figure supplement 3).

In line with this, a positive correlation was observed between the pairwise phylogenetic distances and the normalized NAI IC50 values for each of the tested H6N2s (Figure 2 – figure supplement 4).

We further tested the NAI activity of ferrets’ sera obtained 2 weeks after the prime infection against at least one H6N2 virus from each antigenic cluster. These H6N2 viruses were selected based on their antigenic relatedness as deduced from the ferret immune sera analysis (Figure 2) with 2 representatives from group 1 (Tex12, and Per09) and 1 each from group 2 (Kan17), group 3 (Hel823), and group 4 (Ind11). The normalized data confirmed the antigenic clustering observed in sera after boost (Figure 2 – figure supplement 5).

The NAI titers were also determined in the sera obtained from the 43 groups of NACHO-immunized mice against 7 of the H6N2 viruses (Figure 2 – figure supplement 6a). These H6N2 viruses were selected as described above with the addition of HK14 and Swi13 H6N2s, both belonging to antigenic cluster 1. Serum NAI IC50 values against a given H6N2 virus were normalized against the IC50 of the mouse serum that had the highest NAI titers against the tested H6N2 virus. UPGMA-based clustering showed that the mouse NACHO immune sera displayed a similar recognition pattern as the ferret immune sera that had been generated by H1N2 priming followed by a NACHO boost (Figure 2). For example, HK14 H6N2 NA activity is strongly inhibited by sera from group 1, is poorly inhibited by most of the sera from group 2 and is hardly inhibited by group 3 and 4 sera from both ferrets and mice.

Ind11 H6N2 NA activity, on the other hand, was inhibited by group 1 and 4, and hardly by group 2 and 3 immune sera. Overall, the normalized NAI IC50 values of the ferret and mouse immune sera correlated very well with coefficients values ranging from 0.33 to 0.78 (Figure 2 – figure supplement 6b).

Association analysis reveals amino acid residue substitutions that strongly impact NAI

NA of human influenza viruses is subject to antigenic drift, predominantly driven by the NAI activity that is detectable in the host serum (Air et al., 1985; Yasuhara et al., 2019). There are a total of 102 amino acid differences between the head domains of the NAs that were used to generate the 43 ferret immune sera (Figure 3 - Figure supplement 1). These 102 variable amino acids in the N2 head domain are predominantly exposed on the NA surface (Figure 3 – figure supplement 2).

Association analysis reveals impact of amino acid mutations on NAI.

(a) Association study performed for each H6N2 virus (listed at the bottom) to determine the probability of each variable amino acid (listed on the right) present in the panel of 43 NAs to impact the NAI. The red-to-blue scale represents significance values (S=-log10P) associated with each amino acid substitution. The dark red in the color scale represent all values that are equal or higher than 3 (corresponding to p-values ≤ 0.001). The colour shades on the bottom side indicate the distinct phylogenetic groups. (b) The significant values (p<0.05) were represented on the surface of N2 tetramer (PDB accession number 4H53) for one representative strain of each phylogenetic group with a yellow to red scale of significance (-log10P).

To try to identify which of these amino acid substitutions have a significant impact on NAI, we performed an association analysis for each of the tested H6N2 viruses. For this analysis, the NAI IC50 value was compared between groups of N2 sera according to the presence or absence of a substitution of any of the 102 amino acid position that are different between the panel of 43 tested N2s against which sera had been raised in ferrets. As a result, a p-value is obtained for each position that reflects its possible impact on NAI (Figure 3a). The significance values higher than 1.3 (p<0.05) of representative strains of the 4 antigenic groups were represented on the surface of N2 (Figure 3b). No obvious antigenic patches, however, could be distinguished on the N2 NA surface.

A similar association analysis was performed using the NAI values obtained with the mouse NACHO immune sera. The calculated p-values obtained from the mouse and ferret NAI data were in line with each other (Figure 3 – figure supplement 3).

Residues close to the catalytic pocket are more likely to impact NAI

We next calculated the distance between the Cα atom of each amino acid present in the N2 structure (PDB: 6BR6, from A/Perth/16/2009 NA) to each Cα of the variable amino acids present in the panel. We then determined the correlation of those distances with the p-values that were previously determined in the association analysis. The resulting correlation coefficients (Tau values) indicate a similar pattern among all 27 tested H6N2s (Figure 4a). When visualized on the N2 surface structure, the Tau values revealed a geometrical distribution with the negative correlations being in closer proximity to the catalytic pocket (Figure 4b). Similar data was obtained when mice sera were analyzed this way (Figure 4 – figure supplement 1).

Residues close to the catalytic pocket are more likely to impact NAI.

(a) Significance values (102 variable amino acids) obtained from the association study were correlated to the distance between every amino acid and each of the variable amino acids in the NA structure. The red-to-blue scale represents the Tau values obtained in the Kendall correlation. Amino acids 83 to 469 are shown from top to bottom in the heatmap. The H6N2s used in the NAI panel are listed at the bottom. The grey column indicates the residues that are part of the catalytic pocket (CP) that directly interact with the substrate (R118, D151, R152, R224, E276, R292, R371, and Y406) and the residues that play a key role in stabilizing the catalytic site (E119, R156, W178, S179, D198, I222, E227, H274, E277, N294, and E425). (b) Tau values obtained from Tex12 H6N2 correlation are represented on the surface of the N2 structure of A/RI/5+/1957 (PBD 4H53), the arrow indicates the catalytic pocket in one NA protomer.

Prediction of the N2 antigenicity based on a machine learning method

The antigenic drift of human H3 HA has been extensively studied and modelled (Meyer and Wilke, 2015; Smith et al., 2004; Westgeest et al., 2012; Wu et al., 2020). To model the NA drift observed in our panel of N2 NAs, we first calculated the antigenic distances between the distinct NA sera, using a similar approach as has been used to visualize the HA antigenic map (Cai et al., 2012). Then, using the antigenic distances and a random forest method (RF), a model was built based on amino acid sequence information. A similar model has been used by Yao et al. to predict the antigenicity of H3 HA (Yao et al., 2017). Here, the optimized model was built using a combination of 25 randomly sampled features in 1500 decision trees. The resulting model generates predictive antigenic distances that strongly correlate with the observed values (Figure 5a), with an RMSE/mean=0.207). The RF also enables scoring of amino acid residues according to their importance in the decision trees. One measure of importance can be defined as how early a change in a position is placed in the generated decision trees (mean minimal depth). Another measure of importance is how much the standard error mean increases when a variable is permuted. Finally, the prevalence of a variable amino acid in nodes of the decision tree can also indicate its importance. Those importance scores were represented in a multi-way plot (Figure 5b). The importance scores highlighted similar positions as previously determined in the association study. Finally, we used NA sequences of recent seasonal influenza vaccine recommended H3N2 strains to predict their antigenic distance relative to our panel of N2s, including 3 recent N2 NA components (from A/HongKong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021) that were not present in our data set. The predicted antigenic distances were plotted in the heatmap in a red to blue scale, indicating high to low antigenic distance, respectively (Figure 5c). The RF clearly reproduced the drift observed in AC2 versus AC1. The modelling also indicates that the antigenic distance of the 3 recent N2s to their corresponding group 2 NAs is increasing (represented by the dark blue turning light blue, from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G2) whereas their distance to group 4 NAs is decreasing (dark red becoming yellow from the A/Singapore/infimh-16-0019/2016 towards the A/Darwin/6/2021 in G4).

Prediction of N2 antigenicity based on machine learning.

(a) Predicted vs observed normalized NAIs are shown. (b) Multi-way importance plot indicating the increase in mean standard error, the mean depth, and number of nodes. The top 10 most important amino acid changes are highlighted in blue and annotated with their position in the NA sequence. (c) 43 The heatmap indicates the predicted antigenic distances of N2s from recent H3N2 isolates (at the bottom of the heatmap) relative to the N2s that were used in the NAI breadth panel (listed vertically). The antigenic distances were calculated based on NA protein sequence.

Discussion

The sequence diversity of N2 proteins derived from the 43 selected human H3N2 viruses isolated between 2009 and 2017 allows to classify them in 4 major phylogenetic groups. Group 1 contains 22 N2 NAs, 4 of which were derived from vaccine recommended H3N2 strains: A/Perth/16/2009, A/Texas/50/2012, A/Switzerland/9715293/2013, and A/Hong Kong/4801/2014. Phylogenetic group 2 comprises 14 N2 NAs, one of which is derived from a vaccine H3N2 recommended strain:

A/Singapore/INFIMH-16-0019/2016. The third phylogenetic group contains two NAs that are derived from H3N2 strains that had been isolated in Helsinki in 2013. Remarkably, the NA sequence of these two strains is closely related to that of H3N2s that circulated more than 15 years earlier (e.g. 99% identity with NA of A/Sydney/5/1997). The fourth phylogenetic group is composed of NAs derived from zoonotic cases with swine H3N2 strains isolated between 2012 and 2017. Swine-to-human transmission of H3N2 viruses have occurred multiple times in the USA in 2012 and continue to be detected at low frequency until to date (CDC, 2023; Elbe and Buckland-Merrett, 2017).

We assessed the antigenic diversity of the NA panel by generating 27 recombinant H6N2 viruses and 43 ferret NA immune sera produced by H1N2 priming followed by boosting with homologous recombinant NA (NACHO; n = 41) or by reinfection with H1N2 (A/Estonia/91621/2015 and A/Stockholm/15/2014). Most of the ferrets had detectable serum anti-N2 antibodies by day 14 after infection with H1N2. Anti-NA antibody titers increased after a boost with non-adjuvanted homologous recombinant CHO-produced soluble tetrameric NA suggesting that the use of recombinant NA may boost naturally acquired anti-NA antibody responses in the human population.

The NAI activity of the ferret sera against 27 tested H6N2s reassortants revealed a pattern of inhibition that largely coincides with the phylogenetic distances based on the NA head domain sequences. In the tested panel of H6N2s, at least 4 major antigenic groups were identified. We noted that H6N2 with Kan17 NA had a broader profile of inhibition than other H6N2 viruses from group 2, being also inhibited by serum raised against NA from phylogenetic groups 3 and, to a lesser extent, group 4. Additionally, the H6N2s Wis15 and Ala15 have an antigenic behaviour that is more closely related to antigenic cluster 1, despite belonging to group 2 according to their phylogenetic similarities. Based on the NAI panel, an association analysis was performed to try to identify amino acid changes that affect NAI. It is interesting to note that the amino acid substitutions S245N and S247T, combined with substitution P468H, were identified in this association analysis. Those substitutions were described as major contributors of the NA antigenic drift that occurred in H3N2 in 2014 (Wan et al., 2019). NAs from A/Wisconsin/16/2015 and A/Alaska/251/2015 also have the S245N and S247T substitutions and phylogenetically belong to group 2 NAs. However, those 2 NAs lack the P468H mutation and are antigenically closer to group 1 NAs, confirming the requirement of P468H combined with the S245N and S247T substitutions for this NA antigenic drift (Wan et al., 2019). This observation suggests that NA antigenic drift may sometimes require a combination of amino acid changes rather than single mutations. Wang and colleagues have very elegantly demonstrated that N2 antigenic evolution is constrained by charge balances in an antigenic region, which can partially explain the observation that NA amino acids can undergo co-evolution to result in antigenic drift (Wang et al., 2021). Interestingly, NAs originated from H3N2 variants classified in phylogenetic group 4 were antigenically closer to N2s belonging to phylogenetic group 1. This suggest that the antigenic drift of recent NAs from group 2 increases the antigenic distance from the N2s of swine origin.

Noteworthy, the analysis of the sera obtained from ferrets after infection shows that, despite the lower homologous anti-NA titers, the inhibition pattern evaluated against 5 distinct H6N2s (composed of at least one representative of each antigenic group) is in accordance with the response obtained after boosting with recombinant protein. This suggests that a boost with homologous recombinant soluble tetrameric NA does not alter the breadth of the antibody response induced by H1N2 infection only. Nevertheless, we cannot exclude a degree of increased cross-reactivity of the sera from NACHO278 boosted ferret. It has been reported that repeated H3N2 virus exposure of ferrets affects anti279 hemagglutinin responses (Kosikova et al., 2018)

NAI was also determined against 7 distinct H6N2 reassortant viruses using a panel of 43 anti-N2 NA sera raised in mice by immunization with adjuvanted NACHOs. The observed NAI responses again allowed clustering of the N2 NAs into 4 main antigenic groups, which correlates with data obtained in ferrets. Studies in mice can thus offer a convenient alternative to the use of ferrets in future NA antigenicity studies. This data also supports the hypothesis that there may be universal rules of immunogen antigenicity and immunogenicity (Altman et al., 2015).

The determined KM values of the H6N2 viruses used here, indicates that substrate affinity tends to vary according to the phylogenetic relatedness of N2s. The N-glycosylation at position 245 was previously described to be responsible for reducing N2 affinity for its substrate (Powell and Pekosz, 2020). Unlike group 2, which is primarily characterized by the presence of a N-glycan at position 245, NAs from phylogenetic group 4 have a low KM, indicating an increased affinity for substrate. This result suggests that phylogenetic drift can be accompanied by changes in NA enzymatic properties. Curiously, the Ala15, Wis15, and Kan17, despite having antigenic properties closer to group 1, still have KM values that are comparable to those of group 2 N2 NAs.

After the determination of NA antigenic distances, we used random forest modelling to estimate the impact of amino acid changes on antigenic distances. We obtained a strong correlation of predicted versus observed antigenic distances. However, it is important to note that the training dataset was based on the 102 mutations present in our panel, limiting the predictions for other positions. Furthermore, random forest enabled the classification of the variable amino acids according to importance scores. Also, our modelling does not consider that substitution by other amino acids can have a distinct impact on the antigenic distance. As a consequence, predictions based on the model could underestimate the importance of a particular amino acid residue substitution in some cases. It is important to highlight that co-occurring substitutions in our panel (the ones present in the main branches of the phylogenetic tree) cannot be individually assessed by association analysis or the random forest model. The individual weight of those mutation on NA drift thus remains to be experimentally demonstrated. Interestingly, the modelling could put in evidence the gradual drift of recent N2s which were not present in the dataset (A/Hong Kong/45/2019, A/Tasmania/503/2020, and A/Darwin/6/2021).

It has been reported that an N-glycosylation site at position 329 combined with E344 in NA from human H3N2 viruses from 2007 to 2013 was gradually lost in later H3N2 viruses (Ge et al., 2022). This loss of an N-glycosylation site at position 329 combined with an E344K substitution was associated with a change in NAI reactivity in ferret sera. Three N2 NAs in our panel, derived from A/Victoria/361/2011, A/Hong_Kong/3089/2017, and A/Tennessee/18/2017, lack this N-glycosylation motif. The E344K substitution is present in three other NAs, derived from A/Nagano/2153/2017, A/Minnesota/11/2010, and A/Indiana/08/2011. The importance of those mutations is among the lowest ones predicted by our modeling. However, the differences in NAI reported by Ge et al. are very modest (less than twofold). The experimental variability in our study potentially limits the identification of substitutions with a subtle impact on NAI.

The protective potential of NA antigen against influenza has been underscored by multiple studies. It is likely that NA will be included as a standard component of next generation influenza vaccines. The difficulties of producing stable tetrameric NA may be overcome in the future thanks to protein design or with the advent of mRNA-based vaccines (Catani et al., 2022; Chivukula et al., 2021; Ellis et al., 2022). The antigenic modelling associated with sequence surveillance may increase our predictive ability, highlighting major potential drifts and support NA antigen designs that may elicit broader NAI responses compared to natural NA.

In summary, our study shows that immunization with recombinant N2 NA immunogen can boost anti-NA serum antibody responses in ferrets that were primed by infection with an influenza virus that carries the homologous NA. At least 4 major antigenic groups could be distinguished in N2 NA derived from human H3N2 viruses that were isolated between 2009 to 2017, and those antigenic groups predominantly correspond with the phylogenetic relatedness of the N2 head domains. In addition, the amino acid substitutions in close proximity of the NA catalytic pocket are more likely to affect NAI patterns.

Material and Methods

Rescue of H1N2 and H6N2 reassortants

All reassortant H1N2 and H6N2 viruses included in the study were generated by reverse genetics, with each reassortant expressing the targeted NA antigen, internal genes from A/Puerto Rico/8/1934 H1N1 (PR8), and the HA from PR8 H1N1 or A/mallard/Sweden/81/2002 H6N1, respectively. Briefly, HA and NA segments including non-coding regions were generated by custom gene synthesis (Geneart AG), while PR8 internal segments were derived from a viral isolate as previously published (Marsh and Tannock, 2005). All segments were cloned into a bi-directional transcription plasmid derived from pUC57 (Genscript) including polymerase (Pol) I and Pol II promoters, as previously described (Hoffmann et al., 2000), and the entire set of 8 plasmids was used to transfect 293FT cells (Thermo Fisher Scientific) using Lipofectamine 2000 CD (Thermo Fisher Scientific). Twenty-four hours after transfection, Madin-Darby Canine Kidney Cells (MDCK-) ATL cells (ATCC) were added to the transfected cells in the presence of TPCK-treated trypsin (Sigma) to allow influenza virus propagation. Cell culture supernatants containing influenza virus were harvested 7 days post-MDCK addition and blindly passaged in 8–10-day-old embryonated chicken eggs (Charles River Laboratories, Inc.). Inoculated eggs were incubated at 37°C for 48 h, then cooled to 4°C for 12 h, prior to allantoic fluid harvest and clarification by low-speed centrifugation (3,000 rpm, 20 min). For H6N2 viruses only, high-yield stocks were generated by an additional passage in eggs as described above. Virus titers were determined by plaque assay on MDCK cells.

Design and production of recombinant proteins

The coding information of the NACHO proteins was cloned under the transcriptional control of the CMV promoter in the pCDNA3.4 plasmid with a CD5 secretion signal, an amino-terminal His-tag and the thrombin cleavage signal followed by tetrabrachion-NA. NACHOs were expressed in a mammalian cell culture system as previously described with modifications (Catani et al., 2022).

Animal experiments

The ferret immunization experiment was conducted at Bioqual, Inc., MD (USA). The experiments were approved by the Institutional Animal Care and Use Committee of the Bioqual (reference number 17-021.6). Outbred 18-22 weeks old naïve male and female Fitch ferrets were obtained from Triple F, PA (USA) and housed under specified pathogen-free conditions with food and water ad libitum. Ferrets were primed on day 0 by intranasal challenge with influenza H1N2 reassortant virus (1,000µl/dose, split evenly between nostrils). Four weeks after initial infection (day 28), ferrets were immunized with 45 μg of NACHO (500µl/dose, intramuscularly, split evenly between hind thighs) corresponding to homologous NA used in day 0 prime. Alternatively, and due to low yield of selected recombinant NACHOs, ferrets were boosted on day 28 with 15 μg of homologous NACHO (A/Perth/16/2009nib-64, A/Tasmania/1018/2015, A/Utah/11/2011, and A/West Virginia/17/2012) or were inoculated with the same H1N2 reassortant virus used in the prime (A/Estonia/91621/2015 and A/Stockholm/15/2014). Sera from bleeds two weeks after the prime (day 14) and booster vaccination (day 42) were tested for immunogenicity and seroconversion in serological assays.

The mouse immunization experiments were conducted according to the Belgian legislation (Belgian Law 14/08/1986 and Belgium Royal Decree 06/04/2010) and European legislation on protection of animals used for scientific purposes (EU directives 2010/63/EU and 86/609/EEC). Experimental protocols were all approved by the Ethics Committee of the Vlaams Instituut voor Biotechnologie (VIB), Ghent University, Faculty of Science (EC2019-035). Female BALB/c mice, aged 6–7 weeks, were purchased from Charles River. The mice were housed in a specified pathogen-free animal house with food and water ad libitum. Animals were immunized intramuscularly in the right quadriceps. Immunization was performed with 50 µl containing 1 µg of recombinant protein (tetrabrachion stabilized NAs, referred to as NACHOs). All, except A/West Virginia/17/2012 NACHOs were used as immunogens in mice. All immunizations were adjuvanted with a 1:1 volume of AF03 (25 μl of antigen in PBS + 25 μl of AF03 per dose) (Caillet et al., 2010).

Two or three weeks after the boost immunization, mice were humanely sacrificed with an anaesthetic overdose of Nembutal (6 mg per mouse) and bled retro-orbitally. The obtained blood samples were incubated for 30 min at 37°C to allow clotting, which was followed by centrifugation at 10,500 g for 5 minutes. The supernatant (the serum) was recovered and submitted to a second centrifugation at 10,500 g for 5 minutes. Cleared sera were stored at −20°C before use in serological assays.

ELISA

Anti-tetNA IgG levels in mouse and ferret sera were determined by capture ELISA using tetNAs in PierceTM nickel-coated plates (cat. # 15442). Recombinant proteins were diluted in DPBS (Life technologies cat. # 14040-182) (tetNA was diluted to 0.5 µg/ml). Then, 50 µl of the coating antigen solutions was added to each well and the plate was incubated at room temperature for 1 h on a shaking platform. The wells of the plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and blocked for 1h with 1% BSA (for the mouse sera) or 10% milk (for the ferret sera) in DPBS. After blocking, the wells of the plates were washed once with PBS-T and incubated with a 3-fold serial dilution, starting at a 1/100 dilution, of serum in DPBS, 0.5% BSA, 0.05% tween-20 for 2h at room temperature on a shaking platform. The wells of the plates were then washed 5 times with PBS-T and incubated with a 1:5000 dilution of anti-mouse IgG-HRP (GE healthcare cat. # NA931-1ml), or with a 1:100000 dilution of anti-ferret IgG-HRP (Abcam cat. # 112770) in DPBS, 0.5% BSA, 0.05% Tween-20. The 3,3’,5,5’-tetramethylbenzidine (TMB) substrate (BD cat. # 555214) was added after 3 washes with PBS-T and the reaction was stopped after 5 min by addition of 50 µl of 1 M H2SO4. The optical density (O.D.) in each well was determined at 450 nm and, as a reference, 655 nm using an iMark™ Microplate Absorbance Reader (Bio-rad). The end point titer was determined for each serum sample by scoring the dilution that resulted in an O.D. that was equal to or 2 times higher than the background O.D. obtained from pre-immune control sera dilution series.

Enzyme-linked lectin assay to determine neuraminidase inhibition titers

Fetuin (Sigma cat. # F3385) was diluted into coating buffer (KPL cat. # 50-84-01) to a concentration of 25 µg/ml and 50 µl was added to the wells of Nunc MaxiSorp™ plates (Thermofisher cat. # 44-2404-21), which were incubated overnight at 4°C. The coated plates were then washed 3 times with PBS-T (Sigma cat. # P3563-10PAK) and incubated overnight with 60 μl of a dilution of H6N2 and 60 µl of 2-fold serial dilution of serum, starting at a 1/20 dilution, in sample buffer (1X MES VWR cat. # AAJ61979-AP: 20mM CaCl2, 1% BSA, 0.5% Tween-20). These dilutions of the H6Nx viruses correspond to the 70% maximum activity of NA from the respective viruses as determined in the ELLA assay. Fetuin-coated plates were then washed 3 times with PBS-T and incubated for 1 h with a solution of PNA-HRP (cat. # L6135-1MG, Sigma) at 5 μg/ml in conjugate diluent (MES pH 6.5, 20 mM CaCl2, 1% BSA). The plates were washed 3 times with PBS-T, TMB substrate was added and then the plates were incubated for 5 minutes before the reaction was stopped by the addition of 50 µl of 1 M H2SO4. The optical density was measured at 450 nm and, as a reference, 655 nm in an iMark™ Microplate Absorbance Reader (Bio-rad). Half maximum inhibitory concentrations (IC50) values were determined by non-linear regression analysis (GraphPad Prism software).

NA Enzyme Kinetics Assay

The enzymatic activity of NACHOs and H6N2s was determined with the fluorogenic small substrate 4-methylumbelliferyl)-α-d-N-acetylneuraminic acid (MUNANA) as previously described (Marathe et al., 2013). First H6N2 dilutions were tested to ensure that MUNANA conversion into 4-methylumbelliferone (4-MU) generated fluorescence over time at a linear ratio. Ten microliters of 1 mM of MUNANA (Sigma cat # M8639) in 200 mM sodium acetate buffer (pH 6.5) containing 2 mM CaCl2 and 1% butanol was incubated with 40 µl of PBS solution containing viral dilution. Conversion to 4-MU was monitored every 2 min for 1 h using a BMG Fluostar OPTIMA reader (excitation at 365 nm and emission determined at 450 nm). Second, using the selected dilution of virus, the assay was repeated with dilutions of MUNANA ranging from 2.5 to 100 µM. A standard curve with 800 to 25 pmoles/well of 4-MU (Sigma cat # M1508) was used to extrapolate the molar conversion of MUNANA. KM was calculated using GraphPad Prism version 8.4.3.

Antigenic cartography

Antigenic cartography was performed as described by Smith et al. (Smith et al., 2004) using the R CRAN package (https://CRAN.R-project.org/package=Racmacs). The antigenic map was optimized 5000 times and the best optimization was kept (stress value=1366,47). Uncertainty was measured by bootstrapping 1000 repeats with 100 optimizations per repeat.

Association study

For the association analysis, under the assumption of homogeneous genetic relatedness among the selected H3N2 virus strains, a simple linear regression model for mutation-trait association analysis was fitted to the data for each H6N2 separately. The significance values were calculated by comparing IC50 values of sera that were classified into 2 groups according to the presence or absence of variation in the amino acid of reference Tex12 sequence.

Correlations with distances from the NA catalytic pocket

Tau values were determined by correlating distances with significance values obtained from the association analysis or the importance scores obtained from machine learning. A similar method was described by Meyer et al. to define geometrical constraints of HA evolution (Meyer and Wilke, 2015). Briefly, the relative distances of amino acids were calculated based on the structure of A/Perth/2009 (H3N2) NA (PDB: 6BR6) (Hadházi et al., 2018). The distances were calculated from each Cα to every other Cα. The distances from every residue to the 98 variable residues (4 of the 102 variable amino acids are localized near the end of the stalk and are not resolved in the crystal structure) residues of the that are present in the panel of N2s and the 6BR6 structure were then correlated with the significance values that were obtained in the association analysis, or importance scores obtained in the machine learning. Correlation coefficients were visualized on the structure of N2 NA derived from pandemic A/RI/5+/1957 H2N2 virus (PDB: 4H53)(Vavricka et al., 2013) using Chimera 1.14.

Random Forest algorithm

The Random Forest algorithm was adapted from (Yao et al., 2017). Briefly, the non-conserved positions in the multiple sequence alignment of N2 head domain amino acid sequences of all tested N2s were selected. The feature matrix (X) was created by assessing matches/mismatches on each non-conserved position for each pair of N2 head domain amino acid sequences as: where Xi represents the non-conserved positions in the multiple sequence alignment

m is the number of non-conserved positions (102 in total)

Vi represents the N2 head domain sequence for each tested NA

n is the number of tested sera (43 in total)

0 represents a match, 1 represents a mismatch in the alignment

The response vector (Y) is composed of the antigenic distances for each pair of tested sera. These antigenic distances were determined based on NAI, similarly to the method used to determine HI based immunogenic distances that was described by Cai et al. In summary, the distances between sera was calculated as , which represent average distances between 2 serum samples among all tested H6N2s (Cai et al., 2012). We applied the random forest function from the randomForest package in R to construct the model. We set the bootstrapping (tree) number to 1500 and the number of features to 25 for each tree. Accuracy was determined as the root-mean-square error (RMSE), calculated using the RMSE function from the yardstick package in R.

Additionally, predicted antigenic distances of the out-of-bag samples were compared to the observed distances by fitting a linear model and calculating the R2 value. To calculate the importance of each feature (Xi), the following metrics were calculated using the measure importance function of the randomForestExplainer package in R:

– mean minimal depth: mean of the depths of the node in each tree that splits on this feature and is the closest to the root of the tree (minimal depth). The lower, the more important the feature since a low depth means that many observations are divided into groups based on this feature.

– mse decrease: mean decrease of accuracy in predictions on out of bag samples when the feature is excluded from the model. The higher, the more important the feature.

– number of nodes: total number of nodes in the forest that use the feature for splitting. The higher, the more important the feature.

The data and code used for the generation of the rf model is available at https://github.com/SaelensLAB/RF.